---
title: "Upland United States Southwest Pollen-based Climate Reconstruction"
author: "Andrew Gillreath-Brown"
mainfont: Calibri
output:
  html_document:
    code_folding: show
    keep_md: yes
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
always_allow_html: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, results='hide'}

knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)

# devtools::install()
# devtools::load_all()
library(paleomat)

```

# Introduction

Pollen data can be used to do paleo-temperature reconstructions. However, this type of modeling can be affected by a lot of different aspects, such as paleoecological processes, chronology, and topographic effects on communities and species. 

However, improvements in these techniques, and the increasing breadth of paleoclimatic proxies available have furthered our understanding of the effects of climate-driven variability on past societies. 

This program allows you to reconstruct the climate for multiple locations across North America (when data are sufficient to do). In the program below, you can download fossil and modern data from the [Neotoma Paleoecology Database](https://www.neotomadb.org/), then compile the data using Williams and Shuman (2008) so that there will be columns of taxa with counts, as well as metadata attached to each of those records/rows. Some data in Neotoma overlaps with what was used by Whitmore et al. (2005) in the North American Modern Pollen Database, which can be obtained from one of two sources [the Laboratory for Paleoclimatology and Climatology](http://www.lpc.uottawa.ca/data/modern/) at the University of Ottawa and [the Williams Paleoecology Lab](http://www.geography.wisc.edu/faculty/williams/lab/Downloads.html) at the University of Wisconsin. However, data from the North American Pollen Database is constantly being uploaded to Neotoma, and in some cases corrections are being made to the data too.

# Get fossil, modern, and coretop pollen data.
```{r function_model, echo = FALSE, cache=TRUE, message=FALSE, warning=FALSE}

# Use gpids to get the United States and Canada (or their geopolitical units) in North America. Then get the datasets for the pollen data from each of the gpids.

# Retrieve the GeoPolitical Units table, which has country, state, and county level names with associated IDs. 
gpids <- 
  neotoma::get_table(table.name='GeoPoliticalUnits')

NAID <-  
  gpids %>%
  dplyr::filter(GeoPoliticalName %in% c('United States', 'Canada', 'Mexico'),
                GeoPoliticalUnit == 'country') %$%
  GeoPoliticalID

# BEGIN TESTING for MP_metadata_counts
# AZID <-  
#   gpids %>%
#   dplyr::filter(GeoPoliticalName %in% c('Arizona'),
#                 GeoPoliticalUnit == 'state') %$%
#   GeoPoliticalID
# 
# if(!file.exists(here::here("vignettes/data/AZ_metadata_counts.rds"))){
#   AZ_metadata_counts <- 
#   get_modern_pollen(gpid = AZID) %T>%
#   readr::write_rds(here::here("vignettes/data/AZ_metadata_counts.rds"))
# }
# AZ_metadata_counts <- 
#   readr::read_rds(here::here("vignettes/data/AZ_metadata_counts.rds"))

# END TESTING.

if(!file.exists(here::here("vignettes/data/MP_metadata_counts.rds"))){
  MP_metadata_counts <- 
    get_modern_pollen(gpid = NAID) %T>%
    readr::write_rds(here::here("vignettes/data/MP_metadata_counts.rds"))
}
MP_metadata_counts <- 
  readr::read_rds(here::here("vignettes/data/MP_metadata_counts.rds"))

if(!file.exists(here::here("vignettes/data/NAfossil_metadata_counts.rds"))){
  NAfossil_metadata_counts <- 
    get_fossil_pollen(gpid = NAID) %T>%
    readr::write_rds(here::here("vignettes/data/NAfossil_metadata_counts.rds"))
}
NAfossil_metadata_counts <- 
  readr::read_rds(here::here("vignettes/data/NAfossil_metadata_counts.rds"))

coreTops <- 
  paleomat::get_coreTops() %>% 
  dplyr::rename(
    dataset.id = DatasetID,
    site.id = SiteID, 
    sample.id = SampleID
  )

```

# Load and cleanup modern data (i.e., pollen, climate, and locations)

## Modern Pollen Data from "Core tops" of the Fossil Pollen Dataset in Neotoma

```{r load_modernPollenFromFossil, echo = FALSE, message=FALSE, warning=FALSE, results='hide'}

if(!file.exists(here::here("vignettes/data/NAfossil_metadata_counts.rds"))){
  CT_metadata_counts <- 
  dplyr::left_join(coreTops,
                   NAfossil_metadata_counts,
                   by  = c("dataset.id", "site.id", "sample.id")) %>% 
  dplyr::mutate(type = "core top") %>% 
  dplyr::arrange(dataset.id) %T>%
    readr::write_rds(here::here("vignettes/data/CT_metadata_counts.rds"))
}
CT_metadata_counts <- 
  readr::read_rds(here::here("vignettes/data/CT_metadata_counts.rds"))


```

The Neotoma Modern Pollen Database contains `r nrow(MP_counts)` samples, representing `r ncol(MP_counts)` different pollen taxa.

## Combine Core Top and Modern Pollen Data

```{r Combine_CT_MP, echo = FALSE, message=FALSE, warning=FALSE, results='hide'}

# Now the two dataframes can be combined using the function, then sort the rows by dataset.id. The distinct function is used to make sure there is no duplicate data from combining the two dataframes. Then, we convert all NA values to zero.
MPCT_metadata_counts <- 
  dplyr::bind_rows(MP_metadata_counts, CT_metadata_counts) %>%
  dplyr::arrange(dataset.id) %>%
  dplyr::mutate_at(.vars = dplyr::vars(-dataset.id:-pub_year),
                   .funs = tidyr::replace_na,
                   0) %>%
  dplyr::select(dataset.id:pub_year, 
                sort(tidyselect::peek_vars())) %>%
  dplyr::filter(!is.na(long)) %>% 
  sf::st_as_sf(coords = c("long","lat"), crs = 4326)

# Keep only modern sites in the United States.
# Read in the United States shapefile, which will be used to filter out modern pollen sites that are outside of continental United States.
states <- rgdal::readOGR("vignettes/data/raw_data/statesp010g/statesp010g.shp", layer='statesp010g')
# Transform the states (spatial polygons data frame) to the Coordinate Reference System (CRS) of the PRISM data.
states <- sp::spTransform(states, sp::CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))
# Filter out non-continental states and territories. 
continental.states <- subset(states, NAME != "Alaska" & NAME != "Hawaii" & NAME != "Puerto Rico" & NAME != "U.S. Virgin Islands")
# Convert sp to sf.
continental.states <- sf::st_as_sf(continental.states, crs = 4326)
# Update projection to CRS 4326, which is WGS 84.
continental.states <- sf::st_transform(continental.states, 4326)
# Need to make the geometry valid for the states shapefile.
continental.states <- sf::st_make_valid(continental.states)

# Do an intersection and keep only the sites that are within the continental United States.
sites_in_states <- sf::st_intersection(MPCT_metadata_counts, continental.states)
# Drop the columns from the states shapefile.
MPCT_metadata_counts_final <- dplyr::select (sites_in_states,-c(NAME:PRIM_MILES))



# # Keep only modern samples in specific ecoregions. This is if the reconstructions are limited by their level 1 ecoregions.
# # Read in the Level 1 ecoregions shapefile (https://gaftp.epa.gov/EPADataCommons/ORD/Ecoregions/cec_na/na_cec_eco_l1.zip), which will be used to filter out modern pollen sites that are outside of the given ecoregions.
# ecoregions <- readOGR("/Volumes/PUCK_HD/Users/andrew/Downloads/na_cec_eco_l1/NA_CEC_Eco_Level1.shp")
# # Keep only ecoregions that there are fossil pollen sites in for the UUSS.
# western.ecoregions <- subset(ecoregions, NA_L1CODE == 6 | NA_L1CODE == 10 | NA_L1CODE == 12 | NA_L1CODE == 13)
# # Convert sp to sf.
# western.ecoregions2 <- sf::st_as_sf(western.ecoregions, crs = 4326)
# # Update projection to CRS 4326, which is WGS 84.
# western.ecoregions2 <- sf::st_transform(western.ecoregions2, 4326)
# # Need to make the geometry valid for the states shapefile.
# western.ecoregions2 <- sf::st_make_valid(western.ecoregions2)
# 
# # Keep only ecoregion area that falls within the continental united states.
# ecoregions_in_states <- sf::st_intersection(western.ecoregions2, continental.states)
# 
# # Do an intersection and keep only the sites that are within the selected western ecoregions.
# sites_in_ecoregions <- sf::st_intersection(MPCT_metadata_counts, ecoregions_in_states)
# # Drop the columns from the states shapefile.
# sites_in_ecoregions_final <- dplyr::select (sites_in_ecoregions,-c(NAME:PRIM_MILES))

```

## Remove Coretops data from the fossil pollen dataset
```{r removeCT, echo = FALSE, message=FALSE, warning=FALSE, results='hide'}

fossil_metadata_counts <- 
  dplyr::anti_join(NAfossil_metadata_counts, 
                   coreTops, 
                   by = c("dataset.id", "site.id", "sample.id")) %>%
  sf::st_as_sf(coords = c("long","lat"), crs = 4326)

# # Do an intersection and keep only the sites that are within the continental United States.
# fossil_sites_in_states <- sf::st_intersection(fossil_metadata_counts, continental.states)
# # Drop the columns from the states shapefile.
# fossil_metadata_counts_final <- dplyr::select (fossil_sites_in_states,-c(NAME:PRIM_MILES))

# Limit fossil sites to just the Southwest for reconstruction.
# First, create a bounding box for the upper united states southwest
bb <- c("xmin" = -113, "xmax" = -105, "ymin" = 31.334, "ymax" = 38.09) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  sf::st_as_sf(crs = 4326) %>% 
  sf::st_transform(crs = 4326)

# Do an intersection and keep only the sites that are within the continental United States.
fossil_metadata_counts_final <- sf::st_intersection(fossil_metadata_counts, bb)


fossil_final <-  
  fossil_metadata_counts_final %>%
  dplyr::select(!c(dataset.id, site.id, site.name:pub_year)) %>%
  tibble::as_tibble() %>% 
  dplyr::mutate_all(tidyr::replace_na, 0) %>% 
  dplyr::mutate(dplyr::across(!c(sample.id, geometry), ~ifelse(.x < 0, 0, .x))) %>%
  tidyr::nest(pollen.counts = c(-sample.id, -geometry)) %>%
  sf::st_as_sf() %>%
  dplyr::select(sample.id,
                pollen = pollen.counts)

```

# Load Prism Climate Data

The first step is to get the locations of the Modern Pollen samples. Next, we use a PRISM climate extraction script, which is adapted from [Bocinsky et al. (2016)](https://github.com/bocinsky/Bocinsky_et_al_2016/blob/master/R/Bocinsky_ET_AL_2016_PRISM_EXTRACTOR.R).  

```{r extract_prism_normals, echo = FALSE, warning=FALSE}

#Load prism data. Note: This is only if having trouble installing/loading paleomat package.
# filenames <- list.files("data", pattern="*.rda", full.names=TRUE)
# lapply(filenames,load,.GlobalEnv)
#load(file = "data/prism_1.rda")

# Extract Prism Climate data for modern pollen.
MPCT_climate <- 
  MPCT_metadata_counts_final %>%
  dplyr::select(sample.id) %>% 
  extract_prism_normals() %>%
  sf::st_drop_geometry() %>%
  dplyr::left_join(MPCT_metadata_counts_final, by = "sample.id") %>% 
  dplyr::select(sample.id, 
                prism.normals,
                geometry,
                ABIES:XANTHIUM) %>%
  tidyr::nest(pollen.counts = c(-sample.id, -prism.normals, -geometry)) %>%
  sf::st_as_sf() %>%
  dplyr::select(sample.id, 
                climate = prism.normals,
                pollen = pollen.counts)

# Get the data into the final format used by the palaeoSig and rioja packages.
MPCT_final <- 
  MPCT_climate %>%
  dplyr::mutate(
    # Proportional data is only needed for the significance testing and the proportion are calculated within the palaeoSig::randomTF function.
    # pollen = 
    #   purrr::map(pollen,
    #     function(x){
    #       x/rowSums(x)
    #     }),
    climate = 
      purrr::map(climate,
        function(x){
          x %>%
            dplyr::select(month,
                          ppt,
                          tavg,
                          gdd) %>%
            tidyr::pivot_longer(ppt:gdd,
                                names_to = "element") %>%
            tidyr::unite("element_month", element, month) %>%
            tidyr::pivot_wider(names_from = element_month,
                               values_from = value)
        })
  )


```

## Checking the calibration data set

Need to change this into a function. **

To test the model, I used the `palaeoSig` package's `randomTF()` function, which tests the models against randomly sorted data. If we get a significance for a climate variable, then we know that the model reconstruction is better than random. The model takes the proportion of variance accounted for by the actual data, and then compares it to the proportion of variance accounted for by the randomized data.

The MAT method is testing whether the modern calibration is able to detect signals in each of the climate parameters. The example here uses the entire North American Modern Pollen Database, rather than a targeted data subset.


### MAT - ten closest

```{r MAT_sigTesting_init, results='asis', echo=FALSE, warning=FALSE}

# For some reason this fails consistently when we try to use the weighted MAT (col=2).
palaeoSig::randomTF(spp = 
                      MPCT_final %>%
                      dplyr::select(sample.id, pollen) %>%
                      sf::st_drop_geometry() %>%
                      tidyr::unnest(cols = c(pollen)) %>%
                      tibble::column_to_rownames("sample.id"),
                    env = 
                      MPCT_final %>%
                      dplyr::select(sample.id, climate) %>%
                      sf::st_drop_geometry() %>%
                      tidyr::unnest(cols = c(climate)) %>%
                      tibble::column_to_rownames("sample.id"),
                    fos = 
                      MPCT_final %>%
                      dplyr::select(sample.id, pollen) %>%
                      sf::st_drop_geometry() %>%
                      tidyr::unnest(cols = c(pollen)) %>%
                      tibble::column_to_rownames("sample.id"),
                    n = 99,
                    fun = rioja::MAT,
                    col = 1,
                    k = 10) %$%
  tibble::tibble("variable" = colnames(MPCT_final$climate[[1]]),
                 "% Explained" = EX*100,
                 "p-value" = sig)

```

The variance explained is very low for some variables, although the p value is very low. This highlights that in some cases you may not just want to rely on the p value. This also shows the issue with temporal autocorrelation on temperature and, to some extent also, on precipitation variables.

# Reconstruction Statistics

## Reconstruction Significance

Now we can do a similar test on fossil assemblage reconstructions. This again uses the same `randomTF()` function, but the variance explained and significance will probably change given that we are using a slightly more constrained dataset.

Now, we apply a reconstruction to a real dataset.

#### MAT - ten closest

```{r MAT_sigTesting_fossil, cache=TRUE, results='asis', echo=FALSE, warning=FALSE}

palaeoSig::randomTF(spp = 
                      MPCT_final %>%
                      dplyr::select(sample.id, pollen) %>%
                      sf::st_drop_geometry() %>%
                      tidyr::unnest(cols = c(pollen)) %>%
                      tibble::column_to_rownames("sample.id"),
                    env = 
                      MPCT_final %>%
                      dplyr::select(sample.id, climate) %>%
                      sf::st_drop_geometry() %>%
                      tidyr::unnest(cols = c(climate)) %>%
                      tibble::column_to_rownames("sample.id"),
                    fos = 
                      fossil_final %>%
                      dplyr::select(sample.id, pollen) %>%
                      sf::st_drop_geometry() %>%
                      tidyr::unnest(cols = c(pollen)) %>%
                      tibble::column_to_rownames("sample.id"),
                    n = 99,
                    fun = rioja::MAT,
                    col = 1,
                    k = 10) %$%
  tibble::tibble("variable" = colnames(MPCT_final[[3]][[1]]),
                 "% Explained" = EX*100,
                 "p-value" = sig)


# Old
# This is an altered wrapper for the randomTF function.
# source('../R/sig_test2.R')
# 
# MP_counts_noNAs <- MPCT_metadata_counts_final %>% 
#   dplyr::select(-site.id:-pub_year) %>% 
#   mutate_all(funs(ifelse(is.na(.), 0, .)))
# 
# mat_reconst <- run_tf2(MP_counts_noNAs, fossil = fossil_data, 
#                        climate = Clim_noNAs[,13:24],
#                        func = MAT, col = 2, k = 10)
# 
# mat_reconst[[1]]

# Create a list of all the fossil pollen sites to be reconstructed for the continental United States.
# fossil_sites <- unique(fossil_metadata_counts_final$dataset.id)

# for (i in 1:length(fossil_sites)) {
#   nam <- paste("mat_reconst_", fossil_sites[[i]], sep = "")
#   dat <- run_tf2(pollen = MP_counts_noNAs, fossil = get(fossil_sites[[i]]), 
#                  climate = Temperature,
#                  func = mat, col = 2, k = 10)
#   assign(nam, dat)
# }

```

Again, no significance for the model. This indicates that we are just not able to see a signal within the data, but this is probably due to the calibration dataset being too broad.

## Reconstruction

Once we have validated the methods, we re-run the analyses using MAT.

### Model Summary

```{r clim_reconst, cache=TRUE, results='asis', echo=FALSE, warning=FALSE}

# Now, we carry out the reconstruction for all fossil pollen sites in the southwest US study area.

# Average Temperature
# The reconstructions are in one output file, which has the reconstructed mean temperature (tavg) and associated error for each months.

# First, to save time, check to see if the sites have already been reconstructed, if so then this chunk will not run. If you would like to re-run the reconstructions, then simply delete the reconstruction .csv and it will run.

if (!file.exists("./vignettes/data/reconst_tavg.rds")) {
  mat_models_temp <- purrr::map(
    1:12,
    .f = function(j) {
      rioja::MAT(
        y = dplyr::bind_rows(MPCT_final$pollen),
        x = dplyr::bind_rows(MPCT_final$climate)[[paste0("tavg_", j)]],
        k = 5,
        dist.method = "sq.chord",
        lean = FALSE
      )
    }
  )
  
  reconst_tavg <- mat_models_temp %>%
    purrr::map(
      .f = function(the.model) {
        # Create the MAT reconstruction.
        predict(
          object = the.model,
          newdata = dplyr::bind_rows(fossil_final$pollen) %>%
            as.data.frame() %>%
            set_rownames(fossil_final$sample.id),
          sse = TRUE
        )
      }
    )
  
  readr::write_rds(reconst_tavg, "./vignettes/data/reconst_tavg.rds")
  
} else if (!exists("reconst_tavg")) {
  reconst_tavg <-
    readr::read_rds(here::here("./vignettes/data/reconst_tavg.rds"))
  
} else {
  stop(
    "The SWUS reconstruction is already completed. If you would like to re-run the reconstructions, then delete the file from the directory."
  )
  
}

# GDD
# The reconstructions are in one output file, which has the reconstructed gdd and associated error.

# First, to save time, check to see if the sites have already been reconstructed, if so then this chunk will not run. If you would like to re-run the reconstructions, then simply delete the reconstruction .csv and it will run.

if (!file.exists("./vignettes/data/reconst_gdd.rds")) {
  mat_models_temp <- purrr::map(
    1:12,
    .f = function(j) {
      rioja::MAT(
        y = dplyr::bind_rows(MPCT_final$pollen),
        x = dplyr::bind_rows(MPCT_final$climate)[[paste0("gdd_", j)]],
        k = 5,
        dist.method = "sq.chord",
        lean = FALSE
      )
    }
  )
  
  reconst_gdd <- mat_models_temp %>%
    purrr::map(
      .f = function(the.model) {
        # Create the MAT reconstruction.
        predict(
          object = the.model,
          newdata = dplyr::bind_rows(fossil_final$pollen) %>%
            as.data.frame() %>%
            set_rownames(fossil_final$sample.id),
          sse = TRUE
        )
      }
    )
  
  readr::write_rds(reconst_gdd, "./vignettes/data/reconst_gdd.rds")
  
} else if (!exists("reconst_gdd")) {
  reconst_gdd <-
    readr::read_rds(here::here("./vignettes/data/reconst_gdd.rds"))
  
} else {
  stop(
    "The SWUS reconstruction is already completed. If you would like to re-run the reconstructions, then delete the file from the directory."
  )
  
}


```

### Apply Threshold to Reconstruction

```{r clim_threshold, cache=TRUE, results='asis', echo=FALSE, warning=FALSE}

# Apply a threshold to the reconstruction, so that if an analogue is too dissimilar that it will not be used in the reconstruction. Then, based on the remaining analogues, the final reconstruction and associated error will be re-calculated.

#tavg
if (!file.exists("./vignettes/data/reconst_tavg_threshold.rds")) {
  reconst_tavg_threshold <-
    reconst_tavg %>%
    purrr::map(
      .f = function(the.reconst) {
        apply_threshold(x = the.reconst,
                        qtile = .995)
      }
    )
  
  readr::write_rds(reconst_tavg_threshold,
                   "./vignettes/data/reconst_tavg_threshold.rds")
  
} else if (!exists("reconst_tavg_threshold")) {
  reconst_tavg_threshold <-
    readr::read_rds(here::here("./vignettes/data/reconst_tavg_threshold.rds"))
  
} else {
  stop(
    "The SWUS reconstruction threshold has already been applied. If you would like to re-run the threshold, then delete the file from the directory."
  )
  
}


#gdd
if (!file.exists("./vignettes/data/reconst_gdd_threshold.rds")) {
  reconst_gdd_threshold <-
    reconst_gdd %>%
    purrr::map(
      .f = function(the.reconst) {
        apply_threshold(x = the.reconst,
                        qtile = .995)
      }
    )
  
  readr::write_rds(reconst_gdd_threshold,
                   "./vignettes/data/reconst_gdd_threshold.rds")
  
} else if (!exists("reconst_gdd_threshold")) {
  reconst_gdd_threshold <-
    readr::read_rds(here::here("./vignettes/data/reconst_gdd_threshold.rds"))
  
} else {
  stop(
    "The SWUS reconstruction threshold has already been applied. If you would like to re-run the threshold, then delete the file from the directory."
  )
}
  
```

### Simplify Reconstruction Output

```{r clim_cleanup, cache=TRUE, results='asis', echo=FALSE, warning=FALSE}

# Simplify tavg
# Put the reconstructed climate data into a tibble, using the simplify_reconstruction function. The function takes to parameters, the full reconstruction dataset (reconstruction.data) and the fossil dataset (with metadata and counts) that was used initially loaded that has all of the metadata. The function will select only the metadata and ignore the pollen counts.
if (!file.exists("./vignettes/data/reconst_tavg_final.rds")) {
  reconst_tavg_final <-
    simplify_reconstruction(reconstruction.data = reconst_tavg_threshold,
                            fossil.metadata = fossil_metadata_counts_final) %>%
    # Remove the rows that have NAs for the reconstructed value (i.e., this indicates that there were too few analogues to do a reconstruction). Also, remove rows that do not have an age chronology.
    tidyr::drop_na(age, value)
  
  readr::write_rds(reconst_tavg_final,
                   "./vignettes/data/reconst_tavg_final.rds")
  
} else if (!exists("reconst_tavg_final")) {
  reconst_tavg_final <-
    readr::read_rds(here::here("./vignettes/data/reconst_tavg_final.rds"))
  
  } else {
  stop(
    "The SWUS reconstruction has already been simplified. If you would like to re-run the code, then delete the file from the directory."
  )
  
}


#Simplify GDD

# Put the reconstructed climate data into a tibble, using the simplify_reconstruction function. The function takes to parameters, the full reconstruction dataset (reconstruction.data) and the fossil dataset (with metadata and counts) that was used initially loaded that has all of the metadata. The function will select only the metadata and ignore the pollen counts.
if (!file.exists("./vignettes/data/reconst_gdd_final.rds")) {
  reconst_gdd_final <-
    simplify_reconstruction(reconstruction.data = reconst_gdd_threshold,
                            fossil.metadata = fossil_metadata_counts_final) %>%
    # Remove the rows that have NAs for the reconstructed value (i.e., this indicates that there were too few analogues to do a reconstruction). Also, remove rows that do not have an age chronology.
    tidyr::drop_na(age, value)
  
  readr::write_rds(reconst_gdd_final,
                   "./vignettes/data/reconst_gdd_final.rds")
  
} else if (!exists("reconst_gdd_final")) {
  reconst_gdd_final <-
    readr::read_rds(here::here("./vignettes/data/reconst_gdd_final.rds"))
  
  } else {
  stop(
    "The SWUS reconstruction has already been simplified. If you would like to re-run the code, then delete the file from the directory."
  )
  
}

```

### Mapping via Thin Plate Spline

```{r mapping_tps, cache=TRUE, results='asis', echo=FALSE, warning=FALSE}

# Put lat and long into individual columns.
tavg <- reconst_tavg_final %>%
  tidyr::unnest(geometry) %>%
  dplyr::group_by(sample.id, month) %>%
  dplyr::mutate(col = seq_along(sample.id)) %>% #add a column indicator
  tidyr::spread(key = col, value = geometry) %>%
  dplyr::rename(long = "1", lat = "2")

tavg <- tavg %>% 
  dplyr::mutate(date = paleomat::BPtoBCAD(age)) %>% 
   #dplyr::filter(age < 5950 & age >= 150 & month >= 5 & month <= 9) %>% 
  dplyr::filter(age < 5950 & age >= 150 & month == 7) %>% 
   dplyr::group_by(period=cut(date, breaks=seq(-4000,1900,100)))

tavg$coords <- matrix(c(tavg$long, tavg$lat), byrow=F, ncol=2)

# First, for testing and building the map. I will just use the warmest month (i.e., July), and will only pick 2 centuries to do (i.e., 601-700 and 1201-1300).
# tavg_BC4000 <- tavg %>% 
#      dplyr::filter(period == "(-4e+03,-3.9e+03]")

#600-700 AD
tavg_AD600 <- tavg %>%
  dplyr::filter(period == "(600,700]")

tavg_AD600 <- fields::Tps(tavg_AD600$coords, tavg_AD600$value)
surface(tavg_AD600, nx = 100, ny = 100)

#1200-1300 AD
tavg_AD1200 <- tavg %>%
  dplyr::filter(period == "(1.2e+03,1.3e+03]")

tavg_AD1200 <- fields::Tps(tavg_AD1200$coords, tavg_AD1200$value)
surface(tavg_AD1200)

# Convert reconstruction data to class sf, then get the spatial extent of the fossil pollen sites. 
xys <- sf::st_as_sf(reconst_tavg_final) %>% 
  raster::extent()

# Get extent of fossil pollen sites and make a polygon.
fossilpnts <- reconst_tavg_final %>%
  sf::st_as_sf(crs = 4326)
fossilpolygon <- concaveman(fossilpnts) %>%
  sf::st_as_sf(crs = 4326) %>%
  sf::st_transform(crs = 4326) %>%
  sf:::as_Spatial()

# Create a bounding box for the extent of the sites.
# bb2 <- c("xmin" = xys[1], "xmax" = xys[2], "ymin" = xys[3], "ymax" = xys[4]) %>% 
#   sf::st_bbox() %>% 
#   sf::st_as_sfc() %>% 
#   sf::st_as_sf(crs = 4326) %>% 
#   sf::st_transform(crs = 4326)

#G et prism grid. First, just need one raster layer, which here is just the tmax values for month 7.
prism <- prism_7[2, ]$normal[[1]]
# Do intersection to get just the spatial extent of the fossil pollen sites (via a bounding box or polygon).
UUSS_prism <- raster::mask(prism, fossilpolygon)
# Convert raster layer to a dataframe to put into correct format for the fields package.
UUSS_prism_df <- raster::as.data.frame(UUSS_prism,xy=TRUE) %>% 
  dplyr::select(x, y)

# The prediction locations must be formatted as a matrix, then the predict.Krig function can be called to do the prediction.
UUSS_prism_df$coords <- matrix(c(UUSS_prism_df$x, UUSS_prism_df$y), byrow=F, ncol=2)

# Make predictions onto prism grid (which is limited to the extent of the fossil pollen sites via polygon).
tavg_AD1200.pred <- predict.Krig(tavg_AD1200, UUSS_prism_df$coords)

# Add the predictions (i.e., tavg_AD1200.pred) to the prism grid (i.e., UUSS_prism_df). 
UUSS_prism_df <- UUSS_prism_df %>% 
  dplyr::mutate(pred = tavg_AD1200.pred)


# plot(UUSS_prism_df$coords, pch=20, asp=1,
#  cex=0.8*tavg_AD1200.pred/max(tavg_AD1200.pred),
#  xlab="E", ylab="N",
#  main="Temperature °C")
# 
# ggplot(data = UUSS_prism_df, aes(x, y)) + geom_raster(aes(fill=pred))


# Ordinary Kriging
jura.pred.sp <- tavg_1200
coordinates(jura.pred.sp) <- ~long + lat
v <- variogram(value ~ 1, loc=jura.pred.sp, cutoff = 12)
vmf <- fit.variogram(v, vgm("Sph"))
plot(v, plot.numbers=T, model=vmf,
     main="Fitted variogram model, Jura Co",
     xlab="Separation (km)")

jura.grid.sp <- UUSS_prism_df
coordinates(jura.grid.sp) <- ~x + y
gridded(jura.grid.sp) <- TRUE
k <- krige(value ~ 1, loc=jura.pred.sp, newdata=jura.grid.sp, model=vmf)
```
