---
html_document:
  toc: yes
  toc_depth: 2
  toc_float: yes
  number_sections: no
  code_folding: hide
author: "Andrew Gillreath-Brown"
date: "September 1, 2022"
output:
  html_document:
    df_print: paged
title: "Upland United States Southwest Pollen-based Climate Reconstruction"
mainfont: Calibri
editor_options:
  chunk_output_type: console
keep_md: yes
---

```{r setup, include=FALSE, results='hide'}

knitr::opts_chunk$set(echo = TRUE)

# devtools::install()
# devtools::load_all()
library(paleomat)

```

# Introduction

Pollen data can be used to do paleo-temperature reconstructions. However, this type of modeling can be affected by a lot of different aspects, such as paleoecological processes, chronology, and topographic effects on communities and species. Improvements in these techniques and the increasing breadth of paleoclimatic proxies available, however, have furthered our understanding of the effects of climate-driven variability on past societies.

This program allows you to reconstruct the climate for multiple locations across the southwestern United States. In the program below, you can download fossil and modern data from the [Neotoma Paleoecology Database](https://www.neotomadb.org/), then compile the data using Williams and Shuman (2008), so that there will be columns of taxa with counts and associated metadata attached to each of those records/rows. Some data in Neotoma overlaps with what was used by Whitmore et al. (2005) in the North American Modern Pollen Database, which can be obtained from one of two sources [the Laboratory for Paleoclimatology and Climatology](http://www.lpc.uottawa.ca/data/modern/) at the University of Ottawa and [the Williams Paleoecology Lab](http://www.geography.wisc.edu/faculty/williams/lab/Downloads.html) at the University of Wisconsin. However, data from the North American Pollen Database is constantly being uploaded to Neotoma, and in some cases corrections are being made to the data too. Here, we use the Neotoma package, version 1.7.4.

# Get fossil, modern, and coretop pollen data.
```{r function_model, cache=TRUE, message=FALSE, warning=FALSE}

# Use gpids to get the United States and Canada (or their geopolitical units) in North America. Then, get the datasets for the pollen data from each of the gpids.

# Retrieve the GeoPolitical Units table, which has country, state, and county level names with associated IDs.
gpids <-
  neotoma::get_table(table.name = 'GeoPoliticalUnits')

NAID <-
  gpids %>%
  dplyr::filter(
    GeoPoliticalName %in% c('United States', 'Canada', 'Mexico'),
    GeoPoliticalUnit == 'country'
  ) %$%
  GeoPoliticalID


if (!file.exists(here::here("vignettes/data/pollen_downloads/MP_metadata_counts.rds"))) {
  MP_metadata_counts <-
    paleomat::get_modern_pollen(gpid = NAID) %T>%
    readr::write_rds(here::here("vignettes/data/pollen_downloads/MP_metadata_counts.rds"))
}
MP_metadata_counts <-
  readr::read_rds(here::here("vignettes/data/pollen_downloads/MP_metadata_counts.rds"))

if (!file.exists(here::here("vignettes/data/pollen_downloads/NAfossil_metadata_counts.rds"))) {
  NAfossil_metadata_counts <-
    paleomat::get_fossil_pollen(gpid = NAID) %T>%
    readr::write_rds(here::here(
      "vignettes/data/pollen_downloads/NAfossil_metadata_counts.rds"
    ))
}
NAfossil_metadata_counts <-
  readr::read_rds(here::here(
    "vignettes/data/pollen_downloads/NAfossil_metadata_counts.rds"
  ))

```

# Load and cleanup modern data (i.e., pollen, climate, and locations)

## Modern Pollen Data from "Core tops" of the Fossil Pollen Dataset in Neotoma

```{r load_modernPollenFromFossil, message=FALSE, warning=FALSE, results='hide'}

# This was originally used to pull the core top data; however, the SSL certificate has now expired. The data was saved and can be loaded from the .rds file.
# coreTops <-
#   paleomat::get_coreTops() %>%
#   dplyr::rename(dataset.id = DatasetID,
#                 site.id = SiteID,
#                 sample.id = SampleID)
coreTops <-
  readr::read_rds(here::here("vignettes/data/pollen_downloads/coreTops.rds"))


if (!file.exists(here::here("vignettes/data/pollen_cleaned/CT_metadata_counts.rds"))) {
  CT_metadata_counts <-
    dplyr::left_join(coreTops,
                     NAfossil_metadata_counts,
                     by  = c("dataset.id", "site.id", "sample.id")) %>%
    dplyr::mutate(type = "core top") %>%
    dplyr::arrange(dataset.id) %T>%
    readr::write_rds(here::here("vignettes/data/pollen_cleaned/CT_metadata_counts.rds"))
}
CT_metadata_counts <-
  readr::read_rds(here::here("vignettes/data/pollen_cleaned/CT_metadata_counts.rds"))

```

The Neotoma Modern Pollen Database contains `r nrow(MP_metadata_counts)` samples, representing `r ncol(MP_metadata_counts %>% dplyr::select(ABIES:tail(names(.), 1)))` different pollen taxa.

## Combine Core Top and Modern Pollen Data for Continental United States

```{r Combine_CT_MP, message=FALSE, warning=FALSE, results='hide'}

#Combine the modern and core top data, then keep only sites that fall within the continental United states
if (!file.exists(here::here(
  "vignettes/data/pollen_cleaned/MPCT_metadata_counts_final.rds"
))) {
  # Now, the two dataframes (i.e., the modern data and core top data) can be combined using this function, then sort the rows by dataset.id.
  MPCT_metadata_counts <-
    dplyr::bind_rows(MP_metadata_counts, CT_metadata_counts) %>%
    dplyr::arrange(dataset.id) %>%
    dplyr::mutate(across(-c(dataset.id:pub_year), ~ tidyr::replace_na(.x, 0))) %>%
    dplyr::select(dataset.id:pub_year,
                  sort(tidyselect::peek_vars())) %>%
    dplyr::filter(!is.na(long)) %>%
    sf::st_as_sf(coords = c("long", "lat"), crs = 4326)
  
  # Keep only modern sites in the continental United States.
  
  # Transform the states (spatial polygons data frame) to the Coordinate Reference System (CRS) of the PRISM data.
  states <-
    sp::spTransform(
      paleomat::states,
      sp::CRS(
        "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
      )
    )
  
  # Filter out non-continental states and territories.
  continental.states <-
    subset(
      states,
      NAME != "Alaska" &
        NAME != "Hawaii" &
        NAME != "Puerto Rico" & NAME != "U.S. Virgin Islands"
    )
  
  # Convert sp to sf.
  continental.states <- sf::st_as_sf(continental.states, crs = 4326)
  
  # Update projection to CRS 4326, which is WGS 84.
  continental.states <- sf::st_transform(continental.states, 4326)
  
  # Need to make the geometry valid for the states shapefile.
  continental.states <- sf::st_make_valid(continental.states)
  continental.states_union <- sf::st_union(continental.states)
  
  # Do an intersection and keep only the sites that are within the continental United States.
  buffer <- sf::st_buffer(continental.states_union, 0)
  sites_in_states3 <-
    sf::st_intersection(MPCT_metadata_counts, buffer)
  MPCT_metadata_counts_final <- sites_in_states3
  
  readr::write_rds(
    MPCT_metadata_counts_final,
    here::here(
      "vignettes/data/pollen_cleaned/MPCT_metadata_counts_final.rds"
    )
  )
  
}
MPCT_metadata_counts_final <-
  readr::read_rds(here::here(
    "vignettes/data/pollen_cleaned/MPCT_metadata_counts_final.rds"
  ))


# Create shapefile for only the western  United States, as we will only use modern samples from the western United States. Here, we use the Williams and Shuman 2008 (https://doi.org/10.1016/j.quascirev.2008.01.004) eastern and western splits and use the maximum of all of the splits to designate the extent of the western United States. See extract_WS_pollen_ranges.R for more information.
west_in_states <- paleomat::west_in_states

```

## Remove Coretops data from the fossil pollen dataset
```{r removeCT, message=FALSE, warning=FALSE, results='hide'}

# Remove the core tops from the fossil pollen dataset, and then limit fossil pollen sites to just the southwestern United States.
if (!file.exists(here::here(
  "vignettes/data/pollen_cleaned/fossil_metadata_counts_final2.rds"
))) {
  # Remove the coretop samples from the fossil pollen dataset.
  fossil_metadata_counts <-
    dplyr::anti_join(NAfossil_metadata_counts,
                     coreTops,
                     by = c("dataset.id", "site.id", "sample.id")) %>%
    sf::st_as_sf(coords = c("long", "lat"), crs = 4326)
  
  # Limit fossil pollen sites to just the southwestern United States.
  # First, create a bounding box for the Upper United States Southwest (after Bocinsky and Kohler 2014 - https://doi.org/10.1038/ncomms6618).
  bb <-
    c(
      "xmin" = -113,
      "xmax" = -105,
      "ymin" = 31.334,
      "ymax" = 38.09
    ) %>%
    sf::st_bbox() %>%
    sf::st_as_sfc() %>%
    sf::st_as_sf(crs = 4326) %>%
    sf::st_transform(crs = 4326)
  
  # Do an intersection and keep only the sites that are within the southwestern United States, and remove any samples that do not have an age.
  fossil_metadata_counts_final2 <-
    sf::st_intersection(fossil_metadata_counts, bb) %>%
    dplyr::filter(!is.na(age))
  
  readr::write_rds(
    fossil_metadata_counts_final2,
    here::here(
      "vignettes/data/pollen_cleaned/fossil_metadata_counts_final2.rds"
    )
  )
  
}
fossil_metadata_counts_final2 <-
  readr::read_rds(here::here(
    "vignettes/data/pollen_cleaned/fossil_metadata_counts_final2.rds"
  ))


# Limit samples to post 5500 BP.
if (!file.exists(here::here("data/fossil_west_post5500.rda"))) {
  fossil_west_post5500 <-
    fossil_metadata_counts_final2 %>%
    dplyr::filter(age <= 5500)
  
  save(fossil_west_post5500, file = "data/fossil_west_post5500.rda")
}
fossil_west_post5500 <- paleomat::fossil_west_post5500


# Cleaning up the fossil dataset.
if (!file.exists(here::here("vignettes/data/pollen_cleaned/fossil_final.rds"))) {
  fossil_final <-
    fossil_metadata_counts_final2 %>%
    # Remove any rows that do not have age data.
    dplyr::filter(!is.na(age)) %>%
    dplyr::select(!c(site.id, site.name:pub_year)) %>%
    tibble::as_tibble() %>%
    # For the pollen columns, replace any NA with a 0.
    dplyr::mutate(across(-c(dataset.id, geometry), ~ tidyr::replace_na(.x, 0))) %>%
    # For the pollen columns, replace any negative number with a 0 (in case of any errors).
    dplyr::mutate(dplyr::across(!c(dataset.id, geometry), ~ ifelse(.x < 0, 0, .x))) %>%
    # Nest the pollen data into a nested dataframe.
    tidyr::nest(pollen.counts = -c(dataset.id, geometry)) %>%
    sf::st_as_sf() %>%
    dplyr::select(dataset.id, pollen = pollen.counts) %>%
    # Add column that shows the total number of samples for each core.
    mutate(n_samples = purrr::map_dbl(pollen, nrow))
  
  readr::write_rds(fossil_final,
                   here::here("vignettes/data/pollen_cleaned/fossil_final.rds"))
}
fossil_final <-
  readr::read_rds(here::here("vignettes/data/pollen_cleaned/fossil_final.rds"))

```

## Load Prism Climate Data

The first step is to get the locations of the Modern Pollen samples. Next, we use a PRISM climate extraction script, which is adapted from [Bocinsky et al. (2016)](https://github.com/bocinsky/Bocinsky_et_al_2016/blob/master/R/Bocinsky_ET_AL_2016_PRISM_EXTRACTOR.R).

```{r extract_prism_normals, warning=FALSE}

#Load prism data. Note: This is only if having trouble installing/loading paleomat package.
#filenames <- list.files("data", pattern="*.rda", full.names=TRUE)
#lapply(filenames,load,.GlobalEnv)
#load(file = "data/prism_1.rda")

# Extract Prism climate data for modern pollen sites and limit to the western United States.
if (!file.exists(here::here("vignettes/data/pollen_cleaned/MPCT_climate_final.rds"))) {
  MPCT_climate <-
    MPCT_metadata_counts_final %>%
    dplyr::select(sample.id) %>%
    paleomat::extract_prism_normals() %>%
    sf::st_drop_geometry() %>%
    dplyr::left_join(MPCT_metadata_counts_final, by = "sample.id") %>%
    dplyr::select(sample.id,
                  prism.normals,
                  geometry,
                  ABIES:XANTHIUM) %>%
    tidyr::nest(pollen.counts = -c(sample.id, prism.normals, geometry)) %>%
    sf::st_as_sf() %>%
    dplyr::select(sample.id,
                  climate = prism.normals,
                  pollen = pollen.counts)
  
  # Do an intersection and keep only the sites that are within the western continental United States.
  MPCT_climate_final <-
    sf::st_intersection(MPCT_climate, west_in_states) %>%
    dplyr::select(-FID)
  
  readr::write_rds(
    MPCT_climate_final,
    here::here("vignettes/data/pollen_cleaned/MPCT_climate_final.rds")
  )
}
MPCT_climate_final <-
  readr::read_rds(here::here("vignettes/data/pollen_cleaned/MPCT_climate_final.rds"))

```

## Finalize Modern and Fossil Datasets
```{r finalize-datasets, cache=TRUE, results='asis', warning=FALSE}

# Modern Dataset

# Clean up modern pollen and climate dataframe to prepare for transfer function and reconstruction diagnostics. Here, we transform the pollen data into proportions.
if (!file.exists(here::here("data/MPCT_proportions.rda"))) {
  MPCT_proportions <-
    MPCT_climate_final %>%
    dplyr::mutate(
      pollen =
        purrr::map(pollen,
                   function(x) {
                     x / rowSums(x)
                   }),
      climate =
        purrr::map(climate,
                   function(x) {
                     x %>%
                       dplyr::select(month, tavg)
                   })
    )
  
  save(MPCT_proportions, file = "data/MPCT_proportions.rda")
}
MPCT_proportions <- paleomat::MPCT_proportions

# Put the modern pollen data into a dataframe and make the sample.id as the row names.
spp <- dplyr::bind_rows(MPCT_proportions$pollen) %>%
  dplyr::bind_cols(sample.id = MPCT_proportions$sample.id, .) %>%
  tibble::column_to_rownames("sample.id")

# Put the modern climate data into a dataframe and make the sample.id as the row names.
env <- MPCT_proportions %>%
  dplyr::select(sample.id, climate) %>%
  tidyr::unnest(climate) %>%
  # Pull just the climate (tavg) values as a vector.
  dplyr::pull(tavg)

# Turn climate value (tavg) vector into a named vector with the sample ids.
names(env) <- c(MPCT_proportions$sample.id)


# Fossil Dataset

if (!file.exists(here::here("data/fossil_pollen.rda"))) {
  # Prepare fossil pollen dataset for transfer function and reconstruction diagnostics.
  fossil_final$pollen <-
    purrr::map(fossil_final$pollen, function(x)
      tibble::column_to_rownames(x, "sample.id"))
  
  # Transform fossil pollen data into proportions, as we did with the modern pollen data above.
  fossil_pollen <- fossil_final %>%
    dplyr::mutate(pollen =
                    purrr::map(pollen,
                               function(x) {
                                 x / rowSums(x)
                               }))
  
  save(fossil_pollen, file = "data/fossil_pollen.rda")
}
fossil_pollen <- paleomat::fossil_pollen

```

# Transfer Function and Reconstruction Diagnostics and Reconstruction

Here, we use several transfer function and reconstruction diagnostics, such as squared residual length, the proportion of variance in the fossil data explained by an environmental reconstruction, and the no-analog threshold.

## Squared-residual Length

```{r squared_residual_length, cache=TRUE, results='asis', warning=FALSE}

# Step 1: We use squared residual length. Here, we return the residual length for each sample, and the 0.95 quantile.
if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/rlen_results.rds"
))) {
  rlen_results <-
    purrr::map(
      fossil_pollen$pollen,
      .f = function(x) {
        rlen <- analogue::residLen(spp, env, x, method = "cca")
        
        list(res_lengths = stack(rlen$passive),
             quantile_95 = as.numeric(quantile(rlen$train, probs = 0.95)))
      }
    )
  readr::write_rds(
    rlen_results,
    here::here(
      "vignettes/data/diagnostics_reconstruction/rlen_results.rds"
    )
  )
}
rlen_results <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/rlen_results.rds"
  ))

# Extract the 0.95 quantile baseline to be used for sample removal.
rlen_ninety5quant <- rlen_results[[1]]$quantile_95

# Put the squared residual length and sample.id into one dataframe.
rlen_results <- purrr::map_dfr(rlen_results, `[[`, 1) %>%
  dplyr::rename("Squared_res_lengths" = 1, "sample.id" = 2)

# Extract the sample.ids for samples that will need to be removed since they did not meet the 0.95 threshold.
rlen_removal <-
  as.numeric(levels(rlen_results$sample.id))[rlen_results$Squared_res_lengths > rlen_ninety5quant]

# Results
rlen_full_results1 <- rlen_results %>%
  dplyr::mutate(sample.id = as.integer(levels(sample.id))) %>%
  dplyr::filter(sample.id %in% fossil_west_post5500$sample.id)

if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/rlen_full_results2.rds"
))) {
  rlen_full_results2 <-
    structure(list(res_lengths = rlen_full_results1,
                   quantile_95 = rlen_ninety5quant),
              class = "list")
  
  readr::write_rds(
    rlen_full_results2,
    here::here(
      "vignettes/data/diagnostics_reconstruction/rlen_full_results2.rds"
    )
  )
}
rlen_full_results2 <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/rlen_full_results2.rds"
  ))

fos <- fossil_pollen
fos$pollen <-
  purrr::map(
    fos$pollen,
    ~ .x %>% tibble::rownames_to_column("sample.id") %>% dplyr::filter(!sample.id %in% rlen_removal) %>% tibble::column_to_rownames("sample.id")
  )


# Remove samples that did not meet the 0.95 threshold from the fossil pollen dataset.
if (!file.exists(here::here("vignettes/data/diagnostics_reconstruction/fos2.rds"))) {
  # Remove any sites that have 0 or 1 samples from the fossil pollen dataset.
  fos2 <- fos %>%
    dplyr::select(-n_samples) %>%
    mutate(rows = purrr::map_dbl(pollen, ~ nrow(.x))) %>%
    filter(!rows %in% c(0, 1))
  
  readr::write_rds(fos2,
                   here::here("vignettes/data/diagnostics_reconstruction/fos2.rds"))
}
fos2 <-
  readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/fos2.rds"))

```

## randomTF Testing

```{r randomTF, cache=TRUE, results='asis', warning=FALSE}

# Step 2: We use Random TF to remove sites where tJuly cannot explain at least 90% of the variance.
if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/sig_results2_final.rds"
))) {
  sig_results2_full <-
    purrr::map2(fos2$pollen, fos2$dataset.id, function(x, y) {
      output <- palaeoSig::randomTF(
        spp = spp,
        env = data.frame(tjul = env),
        fos = x,
        n = 10000,
        fun = MAT,
        col = 1
      )
      prop <- output$EX
      names(prop) <- y
      sig_out <- output$sig
      names(sig_out) <- y
      return(list(prop, sig_out))
    })
  
  # Extract the proportion explained variance and the dataset ID.
  sig_results2_proportions <- sapply(sig_results2_full, "[[", 1) %>%
    stack(.) %>%
    dplyr::rename(dataset.id = ind, proportion = values)
  
  # Extract the p-values and the dataset ID.
  sig_results2_pvalues <- sapply(sig_results2_full, "[[", 2) %>%
    stack(.) %>%
    dplyr::rename(dataset.id = ind, pvalue = values)
  
  # Put the proportion of variance and p-value into the same dataframe.
  sig_results2_final <-
    dplyr::left_join(sig_results2_proportions, sig_results2_pvalues, by = "dataset.id") %>%
    dplyr::select(dataset.id, proportion, pvalue) %>%
    dplyr::mutate(dataset.id = as.numeric(levels(dataset.id)))
  
  readr::write_rds(
    sig_results2_final,
    here::here(
      "vignettes/data/diagnostics_reconstruction/sig_results2_final.rds"
    )
  )
}
sig_results2_final <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/sig_results2_final.rds"
  ))

# Pull out the dataset IDs that have p-values less than the 0.10.
only_90b <- sig_results2_final %>%
  dplyr::filter(pvalue <= 0.10) %>%
  dplyr::pull(dataset.id)

# Finally, subset the fossil pollen dataset by keeping only the dataset IDs that passed randomTF.
fos3_90 <- fos %>%
  dplyr::filter(dataset.id %in% only_90b)

```

## MAT Transfer Function and Prediction

```{r MAT_transfer_prediction, cache=TRUE, results='asis', warning=FALSE}

# Step 3: Run MAT and do prediction/reconstruction.

# Remove any columns in the modern pollen dataset that have 0 for the pollen taxa.
pollen_input <- spp %>%
  dplyr::select(which(colSums(.) > 0))

# Remove any columns in the fossil pollen dataset that have 0 for the pollen taxa.
fossil_input <- do.call(rbind, unname(fos3_90$pollen)) %>%
  dplyr::select(which(colSums(.) > 0))

# Keep only columns that are the same between the modern and fossil pollen dataset, as they need to be the same to run MAT.
keep_cols <- colnames(pollen_input[colSums(pollen_input) > 0])
keep_cols <- keep_cols[keep_cols %in% colnames(fossil_input)]

pollen_input <- pollen_input[, keep_cols]
fossil_input <- fossil_input[, keep_cols]

# Save output of MAT if not alaready created.
if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/mat_calibration.rds"
))) {
  # Create the transfer function using mat from the analogue package.
  swap.mat2 <- analogue::mat(pollen_input, env, method = "SQchord")
  
  readr::write_rds(
    swap.mat2,
    here::here(
      "vignettes/data/diagnostics_reconstruction/mat_calibration.rds"
    )
  )
}
swap.mat2 <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/mat_calibration.rds"
  ))

# Here, we run the prediction twice, so that we can extract the minimum dissimilarity from the predict.mat object.

if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/mat_prediction.rds"
))) {
  # Create the transfer function using mat from the analogue package.
  rlgh.mat2 <-
    analogue:::predict.mat(object = swap.mat2,
                           newdata = fossil_input,
                           k = 4)
  
  readr::write_rds(
    rlgh.mat2,
    here::here(
      "vignettes/data/diagnostics_reconstruction/mat_prediction.rds"
    )
  )
}
rlgh.mat2 <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/mat_prediction.rds"
  ))

# Here, we run the prediction again in order to get the bootstrap estimates.
if (!file.exists(
  here::here(
    "vignettes/data/diagnostics_reconstruction/mat_prediction_bootstrap.rds"
  )
)) {
  # Create the transfer function using mat from the analogue package.
  rlgh.mat2b4 <-
    analogue:::predict.mat(
      object = swap.mat2,
      newdata = fossil_input,
      k = 4,
      bootstrap = TRUE
    )
  
  readr::write_rds(
    rlgh.mat2b4,
    here::here(
      "vignettes/data/diagnostics_reconstruction/mat_prediction_bootstrap.rds"
    )
  )
}
rlgh.mat2b4 <-
  readr::read_rds(
    here::here(
      "vignettes/data/diagnostics_reconstruction/mat_prediction_bootstrap.rds"
    )
  )

# Here, we pull out the data from the predict.mat object. First, we pull out the predicted values for the four best analogues. In the analogue package, the analogues have cumulative means, so we only need to extract the fourth record here.
mat_prediction_4 <-
  as.data.frame(rlgh.mat2b4$predictions$model$predicted[4,]) %>%
  tibble::rownames_to_column("sample.id") %>%
  dplyr::rename(value = 2) %>%
  # Pull out the bootstrap predicted values.
  dplyr::left_join(
    .,
    as.data.frame(rlgh.mat2b4[["predictions"]][["bootstrap"]][["predicted"]][, 4]) %>% tibble::rownames_to_column("sample.id") %>%
      dplyr::rename(bootstrap = 2),
    by = "sample.id"
  ) %>%
  # Pull out the bootstrap rmsep.
  dplyr::left_join(
    .,
    as.data.frame(rlgh.mat2b4[["predictions"]][["sample.errors"]][["rmsep"]][, 4]) %>% tibble::rownames_to_column("sample.id") %>%
      dplyr::rename(rmsep = 2),
    by = "sample.id"
  ) %>%
  # Pull out the bootstrap s1 errors.
  dplyr::left_join(
    .,
    as.data.frame(rlgh.mat2b4[["predictions"]][["sample.errors"]][["s1"]][, 4]) %>% tibble::rownames_to_column("sample.id") %>%
      dplyr::rename(s1 = 2),
    by = "sample.id"
  )

```

## No Analog Threshold (or Minimum Dissimilarity)

```{r no_analog, cache=TRUE, results='asis', warning=FALSE}


# Step 4: Remove any fossil pollen samples that have a minimum dissimilarity that is greater than the 5% quantile.
if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/minDC_full_results.rds"
))) {
  # Get a list of the samples that remain in the fossil pollen dataset after rlen and the randomTF removal steps.
  pre_minDC_samples <- rlen_full_results1 %>%
    dplyr::filter(!sample.id %in% rlen_removal) %>%
    dplyr::left_join(
      .,
      fossil_west_post5500 %>%
        dplyr::select(dataset.id, site.id, sample.id),
      by = "sample.id"
    ) %>%
    dplyr::filter(dataset.id %in% only_90b) %>%
    dplyr::pull(sample.id)
  
  # Get the minimum dissimilarity for each sample.
  rlgh.mdc <- analogue::minDC(rlgh.mat2)
  
  # Put the results into a list.
  minDC_results <- stack(rlgh.mdc$minDC) %>%
    dplyr::mutate(ind = as.integer(levels(ind))) %>%
    dplyr::filter(ind %in% pre_minDC_samples) %>%
    dplyr::left_join(
      .,
      fossil_west_post5500 %>%
        dplyr::select(sample.id, dataset.id, site.id, age) %>%
        dplyr::mutate(age = paleomat::BPtoBCAD(age)) %>%
        sf::st_drop_geometry(),
      by = c("ind"  = "sample.id")
    )
  
  minDC_full_results <- structure(list(minDC = minDC_results,
                                       quantiles = rlgh.mdc$quantiles),
                                  class = "list")
  
  readr::write_rds(
    minDC_full_results,
    here::here(
      "vignettes/data/diagnostics_reconstruction/minDC_full_results.rds"
    )
  )
}
minDC_full_results <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/minDC_full_results.rds"
  ))


# Put data into a list. Create list of samples that did and did not meet the threshold, and subset the fossil pollen data to the samples that did pass the minDC test.
if (!file.exists(here::here(
  "vignettes/data/diagnostics_reconstruction/minDC_summary.rds"
))) {
  remove_samples <- as.data.frame(rlgh.mdc$minDC) %>%
    tibble::rownames_to_column("sample.id") %>%
    dplyr::rename(dissimilarity = 2) %>%
    dplyr::filter(dissimilarity > as.numeric(rlgh.mat2$quantiles["5%"]))
  
  
  mat_prediction_cleaned <- mat_prediction_4 %>%
    dplyr::filter(!sample.id %in% remove_samples$sample.id)
  
  # Final Sample set.
  final_approved_samples <- minDC_full_results$minDC %>%
    dplyr::filter(!ind %in% remove_samples$sample.id) %>%
    dplyr::pull(ind)
  
  minDC_summary <- list(
    remove_samples = remove_samples,
    mat_prediction_cleaned = mat_prediction_cleaned,
    final_approved_samples = final_approved_samples
  )
  
  readr::write_rds(
    minDC_summary,
    here::here(
      "vignettes/data/diagnostics_reconstruction/minDC_summary.rds"
    )
  )
}
minDC_summary <-
  readr::read_rds(here::here(
    "vignettes/data/diagnostics_reconstruction/minDC_summary.rds"
  ))

# Need to update age models for some fossil pollen sites that have not been updated with the Bacon age models. Here, we read in the updated age models from individual .csv files.
updated_age_models <-
  list.files("vignettes/data/updated_age_models", full.names = TRUE) %>%
  purrr::map_dfr(read.csv)

# Clean up fossil pollen data to create the final dataset of samples.
if (!file.exists(here::here("vignettes/data/diagnostics_reconstruction/fos_90.rds"))) {
  fos_90 <- mat_prediction_cleaned %>%
    dplyr::mutate(sample.id = as.integer(sample.id)) %>%
    dplyr::filter(sample.id %in% final_approved_samples) %>%
    dplyr::left_join(.,
                     fossil_metadata_counts_final %>%
                       dplyr::select(dataset.id:pub_year),
                     by = "sample.id") %>%
    dplyr::filter(age <= 5500) %>%
    dplyr::left_join(updated_age_models,
                     by = c("site.id", "dataset.id", "sample.id")) %>%
    dplyr::mutate(age = dplyr::coalesce(age.y, age.x)) %>%
    dplyr::select(-c(age.y, age.x)) %>%
    dplyr::mutate(age = round(age)) %>%
    sf::st_as_sf()
  
  readr::write_rds(fos_90,
                   here::here("vignettes/data/diagnostics_reconstruction/fos_90.rds"))
}
fos_90 <-
  readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/fos_90.rds"))







# COMBO #3 - ONLY NO ANALOGUE REMOVAL.

# fos <- fossil_pollen
# fos2 <- fos %>% 
#   dplyr::select(-n_samples) %>% 
#   mutate(rows = map_dbl(pollen, ~ nrow(.x))) %>% 
#   filter(!rows %in% c(0, 1))
# 
# pollen_input <- spp %>% 
#   dplyr::select(which(colSums(.) > 0))
# 
# fossil_input <- do.call(rbind, unname(fos2$pollen)) %>% 
#   dplyr::select(which(colSums(.) > 0))
# 
# keep_cols <- colnames(pollen_input[colSums(pollen_input) > 0])
# keep_cols <- keep_cols[keep_cols %in% colnames(fossil_input)]
# 
# pollen_input <- pollen_input[, keep_cols]
# fossil_input <- fossil_input[, keep_cols]
# 
# 
# swap.mat2 <- mat(pollen_input, env, method = "SQchord")
# 
# rlgh.mat2 <- predict(object = swap.mat2, newdata = fossil_input)
# 
# mat_prediction <- as.data.frame(rlgh.mat2$predictions$model$predicted[4,]) %>%
#   tibble::rownames_to_column("sample.id") %>%
#   rename(value = 2)
# 
# rlgh.mdc <- minDC(rlgh.mat2)
# 
# remove_samples <- as.data.frame(rlgh.mdc$minDC) %>%
#   tibble::rownames_to_column("sample.id") %>%
#   rename(dissimilarity = 2) %>%
#   filter(dissimilarity > as.numeric(rlgh.mat2$quantiles["5%"]))
# 
# mat_prediction_cleaned <- mat_prediction %>%
#   filter(!sample.id %in% remove_samples$sample.id)
# 
# fos_90 <- mat_prediction_cleaned %>% 
#   mutate(sample.id = as.integer(sample.id)) %>% 
#   dplyr::left_join(., fossil_metadata_counts_final %>% 
#   dplyr::select(dataset.id:pub_year), by = "sample.id") %>% 
#   #dplyr::mutate(error = new_output_recont_fos3_90$SEP.boot[, "MAT"]) %>% 
#   filter(age <= 5500) %>% 
#   dplyr::left_join(updated_age_models, by = c("site.id", "dataset.id", "sample.id")) %>%
#   dplyr::mutate(age = dplyr::coalesce(age.y, age.x)) %>%
#   dplyr::select(-age.y, -age.x) %>%
#   dplyr::mutate(age = round(age)) %>% 
#   # filter(site.id == 246) %>% 
#   sf::st_as_sf()
# 
# write.csv(fos_90, "Tim_NoAnalogue_Only_1best_analogs.csv", row.names = FALSE)

```

# Anomaly Calculation

```{r anomaly_calculation, cache=TRUE, results='asis', warning=FALSE}

# Get the PRISM modern July temperature dataset.
temp.raster <-
  raster::raster(here::here("vignettes/data/spatial/temp.raster.tif"))

# Extract the modern mean temperature at each fossil pollen site, and then calculate the anomalies.
if (!file.exists(
  here::here(
    "vignettes/data/diagnostics_reconstruction/final_paleomat_dataset3.rds"
  )
)) {
  # Put lat and long into individual columns.
  tavg_new_90 <- fos_90 %>%
    dplyr::mutate(long = unlist(purrr::map(.$geometry, 1)),
                  lat = unlist(purrr::map(.$geometry, 2))) %>%
    dplyr::select(-geometry) %>%
    sf::st_drop_geometry() %>%
    dplyr::mutate(date = paleomat::BPtoBCAD(age)) %>%
    dplyr::filter(age <= 5500) %>%
    dplyr::group_by(period = cut(date, breaks = seq(-5500, 2100, 100))) %>%
    dplyr::mutate(mid_period = paleomat::get_midpoints(period))
  
  modern_temperature2 <-
    as.data.frame(
      raster::extract(
        x = temp.raster,
        y = tavg_new_90 %>% dplyr::ungroup() %>%
          dplyr::select(long, lat)
      )
    ) %>%
    dplyr::rename(modern = 1)
  
  # Bind the modern data to fossil pollen dataset.
  site.preds.unlisted <-
    dplyr::bind_cols(tavg_new_90, modern = modern_temperature$modern)
  
  # Calculate the temperature anomaly and the bootstrapped temperature anomaly.
  site.preds.unlisted.anom <- site.preds.unlisted %>%
    dplyr::rowwise() %>%
    dplyr::mutate(anom = value - modern) %>%
    dplyr::mutate(anom_bootstrap = bootstrap - modern)
  
  readr::write_rds(
    site.preds.unlisted.anom,
    here::here(
      "vignettes/data/diagnostics_reconstruction/final_paleomat_dataset3.rds"
    )
  )
}
site.preds.unlisted.anom <-
  readr::read_rds(
    here::here(
      "vignettes/data/diagnostics_reconstruction/final_paleomat_dataset3.rds"
    )
  )

```

For figures and tables in the journal article, please see "Paleomat Figures.Rmd".

