---
author: "Andrew Gillreath-Brown"
date: "`r format(Sys.time(), '%d %B %Y')`"
title: "paleoMAT: A Low-Frequency Summer Temperature Reconstruction for the United States Southwest, 3000 BC – AD 2000"
html_document:
  toc: yes
  toc_depth: 2
  toc_float: yes
  number_sections: no
  code_folding: hide
output:
  html_document:
    df_print: paged
mainfont: Calibri
editor_options:
  chunk_output_type: console
keep_md: yes
---

```{r setup, include = FALSE, results = 'hide'}

knitr::opts_chunk$set(echo = TRUE)

# If you have cloned the repository, then you can also use devtools::install() to install the package.
# devtools::install()
# devtools::load_all()
library(paleomat)

```

# Introduction

Pollen data can be used to do paleo-temperature reconstructions. However, this type of modeling can be affected by a lot of different aspects, such as paleoecological processes, chronology, and topographic effects on communities and species. Improvements in these techniques and the increasing breadth of paleoclimatic proxies available, however, have furthered our understanding of the effects of climate-driven variability on past societies.

This program allows you to reconstruct the climate for multiple locations across the southwestern United States. In the program below, you can download fossil and modern data from the [Neotoma Paleoecology Database](https://www.neotomadb.org/), then compile the data using Williams and Shuman (2008), so that there will be columns of taxa with counts and associated metadata attached to each of those records/rows. Some data in Neotoma overlaps with what was used by Whitmore et al. (2005) in the North American Modern Pollen Database, which can be obtained from one of two sources [the Laboratory for Paleoclimatology and Climatology](http://www.lpc.uottawa.ca/data/modern/) at the University of Ottawa and [the Williams Paleoecology Lab](http://www.geography.wisc.edu/faculty/williams/lab/Downloads.html) at the University of Wisconsin. However, data from the North American Pollen Database is constantly being uploaded to Neotoma, and in some cases corrections are being made to the data too. Here, we use the Neotoma package, version 1.7.4.

**Note**: While running the code below, if you have trouble with `paleomat` datasets (e.g., `paleomat::coreTops`), then you can use `here::here` with `data` as the root folder, then the variable name with `.rda`, such as `load(file = here::here("data/coreTops.rda"))`, to load a dataset. Additionally, we use `if` statements in a number of places throughout the markdown to save time, as some items can take a while to run. However, if you have cloned the repository, then you can run the interior contents of those if statements to overwrite the variables.

# Download Fossil and Modern Pollen Data

```{r function_model, cache = TRUE, message = FALSE, warning = FALSE}

# Use gpids to get the United States and Canada (or their geopolitical units) in North America. Then, get the datasets for the pollen data from each of the gpids. Here, to save time, we read in the already downloaded data from Neotoma, as it takes a while to run for the 3 countries. Data downloaded on October 23, 2021.

# Retrieve the GeoPolitical Units table, which has country, state, and county level names with associated IDs. We skip this step if the fossil pollen and modern pollen datasets are present.
if (!file.exists(here::here("data/MP_metadata_counts.rda")) | !file.exists(here::here("data/NAfossil_metadata_counts.rda"))) {
gpids <-
  neotoma::get_table(table.name = 'GeoPoliticalUnits')

NAID <-
  gpids %>%
  dplyr::filter(
    GeoPoliticalName %in% c('United States', 'Canada', 'Mexico'),
    GeoPoliticalUnit == 'country'
  ) %$%
  GeoPoliticalID
}

# Get the modern pollen dataset.
if (!file.exists(here::here("data/MP_metadata_counts.rda"))) {
  MP_metadata_counts <-
    paleomat::get_modern_pollen(gpid = NAID)
  
  save(MP_metadata_counts,
       file = here::here("data/MP_metadata_counts.rda"))
}
MP_metadata_counts <- paleomat::MP_metadata_counts

# Get the fossil pollen dataset.
if (!file.exists(here::here("data/NAfossil_metadata_counts.rda"))) {
  NAfossil_metadata_counts <-
    paleomat::get_fossil_pollen(gpid = NAID)
  
  save(NAfossil_metadata_counts,
       file = here::here("data/NAfossil_metadata_counts.rda"))
}
NAfossil_metadata_counts <- paleomat::NAfossil_metadata_counts

```

The Neotoma Modern Pollen Database contains `r nrow(MP_metadata_counts)` samples for the United States, Canada, and Mexico, representing `r ncol(MP_metadata_counts %>% dplyr::select(ABIES:tail(names(.), 1)))` different pollen taxa.

# Data Cleaning for Fossil and Modern Pollen and Climate Data

## Load "Core Tops" Data from the Fossil Pollen Dataset

```{r load_coreTops, message = FALSE, warning = FALSE, results = 'hide'}

# Get core tops from the fossil pollen dataset and add to the modern pollen dataset. Data was originally downloaded using Tilia with a function created by Eric Grimm. The SSL certificate has now expired, but the sample numbers, dataset ID, and site ID core tops can be loaded using paleomat::coreTops.
if (!file.exists(here::here("vignettes/data/pollen_cleaned/CT_metadata_counts.rds"))) {
  CT_metadata_counts <-
    dplyr::left_join(paleomat::coreTops,
                     NAfossil_metadata_counts,
                     by  = c("dataset.id", "site.id", "sample.id")) %>%
    dplyr::mutate(type = "core top") %>%
    dplyr::arrange(dataset.id) %T>%
    readr::write_rds(here::here("vignettes/data/pollen_cleaned/CT_metadata_counts.rds"))
}
CT_metadata_counts <-
  readr::read_rds(here::here("vignettes/data/pollen_cleaned/CT_metadata_counts.rds"))

```

## Combine Core Top and Modern Pollen Data for Continental United States

```{r Combine_CT_MP, message = FALSE, warning = FALSE, results = 'hide'}

#Combine the modern and core top data, then keep only sites that fall within the continental United states.

# If you do not want to run this code chunk to create MPCT_metadata_counts_final, then you can uncomment and run the following line:
# MPCT_metadata_counts_final <-
#   readr::read_rds(here::here("vignettes/data/pollen_cleaned/MPCT_metadata_counts_final.rds"))

# Now, the two dataframes (i.e., the modern data and core top data) can be combined using this function, then sort the rows by dataset.id.
MPCT_metadata_counts <-
  dplyr::bind_rows(MP_metadata_counts, CT_metadata_counts) %>%
  dplyr::arrange(dataset.id) %>%
  dplyr::mutate(across(-c(dataset.id:pub_year), ~ tidyr::replace_na(.x, 0))) %>%
  dplyr::select(dataset.id:pub_year,
                sort(tidyselect::peek_vars())) %>%
  dplyr::filter(!is.na(long)) %>%
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326)

# Keep only modern sites in the continental United States.

# Transform the states (spatial polygons data frame) to the Coordinate Reference System (CRS) of the PRISM data.
states <-
  sp::spTransform(
    paleomat::states,
    sp::CRS(
      "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"
    )
  )

# Filter out non-continental states and territories.
continental.states <-
  subset(
    states,
    NAME != "Alaska" &
      NAME != "Hawaii" &
      NAME != "Puerto Rico" & NAME != "U.S. Virgin Islands"
  )

# Convert sp to sf.
continental.states <- sf::st_as_sf(continental.states, crs = 4326)

# Update projection to CRS 4326, which is WGS 84.
continental.states <- sf::st_transform(continental.states, 4326)

# Need to make the geometry valid for the states shapefile.
continental.states <- sf::st_make_valid(continental.states)
continental.states_union <- sf::st_union(continental.states)

# Do an intersection and keep only the sites that are within the continental United States.
buffer <- sf::st_buffer(continental.states_union, 0)

MPCT_metadata_counts_final <-
  suppressWarnings(sf::st_intersection(MPCT_metadata_counts, buffer))

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   MPCT_metadata_counts_final,
#   here::here("vignettes/data/pollen_cleaned/MPCT_metadata_counts_final.rds"))

#------End for MPCT_metadata_counts_final--------#


```

## Remove Coretops Data from the Fossil Pollen Dataset

```{r removeCT, message = FALSE, warning = FALSE, results = 'hide'}

# Remove the core tops from the fossil pollen dataset, and then limit fossil pollen sites to just the southwestern United States.

# If you do not want to run the code for creating fossil_metadata_counts_final2, then you can uncomment and run the following line:
# fossil_metadata_counts_final2 <- readr::read_rds(here::here("vignettes/data/pollen_cleaned/fossil_metadata_counts_final2.rds"))

# Remove the coretop samples from the fossil pollen dataset.
fossil_metadata_counts <-
  dplyr::anti_join(NAfossil_metadata_counts,
                   coreTops,
                   by = c("dataset.id", "site.id", "sample.id")) %>%
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326)

# Limit fossil pollen sites to just the southwestern United States.
# First, create a bounding box for the Upper United States Southwest (after Bocinsky and Kohler 2014 - https://doi.org/10.1038/ncomms6618).
bb <-
  c(
    "xmin" = -113,
    "xmax" = -105,
    "ymin" = 31.334,
    "ymax" = 38.09
  ) %>%
  sf::st_bbox() %>%
  sf::st_as_sfc() %>%
  sf::st_as_sf(crs = 4326) %>%
  sf::st_transform(crs = 4326)

# Do an intersection and keep only the sites that are within the southwestern United States. 
fossil_metadata_counts_final2 <-
  suppressWarnings(sf::st_intersection(fossil_metadata_counts, bb))

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   fossil_metadata_counts_final2,
#   here::here("vignettes/data/pollen_cleaned/fossil_metadata_counts_final2.rds"))

#------End for fossil_metadata_counts_final2--------#


```

## Add Bacon Age-depth Models to Fossil Pollen Dataset

```{r age_models, message = FALSE, warning = FALSE, results = 'hide'}

# If you do not want to run the code for creating fossil_metadata_counts_final3, then you can uncomment and run the following line:
# fossil_metadata_counts_final3 <- readr::read_rds(here::here("vignettes/data/pollen_cleaned/fossil_metadata_counts_final3.rds"))

# Keep only sites that have new age models. Most sites that are removed from fossil_metadata_counts_final2 did not have ages, and were primarily archaeological sites with one sample.
fossil_metadata_counts_final3 <- fossil_metadata_counts_final2 %>% 
  dplyr::filter(dataset.id %in% paleomat::bacon_age_models$dataset.id) %>% 
  dplyr::left_join(., paleomat::bacon_age_models, by = c("dataset.id", "site.id", "depth")) %>% 
  dplyr::relocate(min, max, median, mean, .after = age) %>% 
  dplyr::select(-site.name.y) %>% 
  dplyr::rename(site.name = site.name.x)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   fossil_metadata_counts_final3,
#   here::here("vignettes/data/pollen_cleaned/fossil_metadata_counts_final3.rds"))

```

## Apply Quality Control Filters to the Fossil Pollen Dataset

```{r clean_fossil_dataset, message = FALSE, warning = FALSE, results = 'hide'}

# If you do not want to run the code for creating fossil_metadata_counts_final4, then you can uncomment and run the following line:
# fossil_metadata_counts_final4 <- readr::read_rds(here::here("vignettes/data/pollen_cleaned/fossil_metadata_counts_final4.rds"))

# Apply quality control filters to the fossil pollen dataset.
fossil_metadata_counts_final4 <- fossil_metadata_counts_final3 %>% 
  # Calculate the total pollen grains identified in each sample. 
  dplyr::mutate(total_grains = rowSums(dplyr::select(sf::st_drop_geometry(.), ABIES:XANTHIUM))) %>% 
  # A sample must have a minimum of 200 counted pollen grains to be included in the dataset.
  dplyr::filter(total_grains >= 200) %>% 
  # Convert to proportions.
  dplyr::mutate(dplyr::across(ABIES:XANTHIUM, ~.x / total_grains)) %>%
  # Replace observations with < 1% abundance with 0.
  dplyr::mutate(dplyr::across(ABIES:XANTHIUM, ~dplyr::if_else(.x < .005, 0, .x)))

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   fossil_metadata_counts_final4,
#   here::here("vignettes/data/pollen_cleaned/fossil_metadata_counts_final4.rds"))

# Remove taxa that do not occur in at least 20 samples. Retrieve column names to remove.
rm_taxa <- fossil_metadata_counts_final4 %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(ABIES:XANTHIUM) %>% 
  dplyr::select_if(~!(is.numeric(.) && (sum(.x > 0) > 20))) %>% 
  colnames()

fossil_metadata_counts_final5 <- fossil_metadata_counts_final4 %>% 
  # Remove taxa columns that did not have at least 20 samples.
  dplyr::select(-c(dplyr::all_of(rm_taxa))) %>% 
  # Select sites that have 5 or more unique taxa.
  dplyr::mutate(taxa_count = rowSums(!dplyr::near(dplyr::select(sf::st_drop_geometry(.), ABIES:THALICTRUM), 0))) %>%
  dplyr::filter(taxa_count >= 5) %>%
  dplyr::select(-taxa_count, -total_grains)

# Limit samples to post 6000 BP.
fossil_west_post5500 <- fossil_metadata_counts_final5 %>% 
  dplyr::filter(max <= 6000)

# If you want to re-save fossil_west_post5500.
#save(fossil_west_post5500, file = here::here("data/fossil_west_post5500.rda"))

#------End for fossil_west_post5500--------#


# Cleaning up the fossil dataset.

# If you do not want to run the code for creating fossil_final, then you can uncomment and run the following line:
# fossil_final <-
#   readr::read_rds(here::here("vignettes/data/pollen_cleaned/fossil_final.rds"))

fossil_final <-
  fossil_west_post5500 %>%
  # Remove any rows that do not have age data.
  dplyr::filter(!is.na(age)) %>%
  dplyr::select(!c(site.id, site.name:pub_year)) %>%
  tibble::as_tibble() %>%
  # Nest the pollen data into a nested dataframe.
  tidyr::nest(pollen.counts = -c(dataset.id, geometry)) %>%
  sf::st_as_sf() %>%
  dplyr::select(dataset.id, pollen = pollen.counts) %>%
  # Add column that shows the total number of samples for each core.
  dplyr::mutate(n_samples = purrr::map_dbl(pollen, nrow))

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(fossil_final,
#                  here::here("vignettes/data/pollen_cleaned/fossil_final.rds"))


#------End for fossil_final--------#

```

## Load Prism Climate Data

The first step is to get the locations of the Modern Pollen samples. Next, we use a PRISM climate extraction script, which is adapted from [Bocinsky et al. (2016)](https://github.com/bocinsky/Bocinsky_et_al_2016/blob/master/R/Bocinsky_ET_AL_2016_PRISM_EXTRACTOR.R).

```{r extract_prism_normals, warning = FALSE}

# Extract Prism climate data for modern pollen sites and limit to the western United States.

# If you do not want to run the code for creating MPCT_climate_final, then you can uncomment and run the following line:
# MPCT_climate_final <-
#   readr::read_rds(here::here("vignettes/data/pollen_cleaned/MPCT_climate_final.rds"))

MPCT_climate <-
  MPCT_metadata_counts_final %>%
  dplyr::select(sample.id) %>%
  paleomat::extract_prism_normals() %>%
  sf::st_drop_geometry() %>%
  dplyr::left_join(MPCT_metadata_counts_final, by = "sample.id") %>%
  dplyr::select(sample.id,
                elev,
                prism.normals,
                geometry,
                ABIES:XANTHIUM) %>% 
  sf::st_as_sf()

# Do an intersection and keep only the sites that are within the western continental United States.
MPCT_climate_final <-
  suppressWarnings(sf::st_intersection(MPCT_climate, paleomat::west_in_states)) %>%
  dplyr::select(-FID)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   MPCT_climate_final,
#   here::here("vignettes/data/pollen_cleaned/MPCT_climate_final.rds")
# )

```

# Apply Quality Control Filters to Modern Pollen Dataset

```{r modern_quality control, warning = FALSE}

MPCT_climate_final2 <- MPCT_climate_final %>% 
  dplyr::mutate(total_grains = rowSums(dplyr::select(sf::st_drop_geometry(.), ABIES:XANTHIUM))) %>% 
  # A sample must have a minimum of 200 counted pollen grains to be included in the dataset.
  dplyr::filter(total_grains >= 200) %>% 
  dplyr::mutate(dplyr::across(ABIES:XANTHIUM, ~.x / total_grains)) %>%
  # Replace observations with < 1% abundance with 0.
  dplyr::mutate(dplyr::across(ABIES:XANTHIUM, ~dplyr::if_else(.x < .005, 0, .x)))

# Remove taxa that do not occur in at least 20 samples. Retrieve column names to remove.
rm_taxa <- MPCT_climate_final2 %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(ABIES:XANTHIUM) %>% 
  dplyr::select_if(~!(is.numeric(.) && (sum(.x > 0) > 20))) %>% 
  colnames()

MP_coords <- dplyr::bind_rows(MP_metadata_counts, CT_metadata_counts) %>% 
  dplyr::select(sample.id, long, lat)

MPCT_climate_final3 <- MPCT_climate_final2 %>% 
  dplyr::select(-c(dplyr::all_of(rm_taxa))) %>% 
  # Select sites that have 5 or more unique taxa.
  dplyr::mutate(taxa_count = rowSums(!dplyr::near(dplyr::select(sf::st_drop_geometry(.), ABIES:ULMUS), 0))) %>%
  dplyr::filter(taxa_count >= 5) %>%
  dplyr::select(-taxa_count, -total_grains) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::left_join(., MP_coords, by = "sample.id") %>% 
  tidyr::nest(pollen.counts = -c(sample.id, elev, prism.normals, long, lat)) %>%
  dplyr::select(sample.id,
                elev,
                climate = prism.normals,
                pollen = pollen.counts,
                long, 
                lat) %>% 
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326)

# Check that the recorded elevation from Neotoma is not too different from elevation (using the elevatr package) of the location.
df_elev_epqs <- elevatr::get_elev_point(MPCT_climate_final3 %>% dplyr::select(-c(climate, pollen)), prj = 4326, src = "aws") %>% 
  # Calculate the absolute difference between the elevations.
  dplyr::mutate(diff = abs(elevation - elev))

# Remove samples that do not match the DEM.
mismatch <- df_elev_epqs %>% 
  dplyr::filter(diff > 1000) %>% 
  dplyr::pull(sample.id)

MPCT_climate_final3 <- MPCT_climate_final3 %>% 
  dplyr::filter(!sample.id %in% mismatch) %>% 
  dplyr::select(-elev)

```

## Finalize Modern and Fossil Pollen Datasets
```{r finalize-datasets, cache = TRUE, results = 'asis', warning = FALSE}

# Modern Dataset

# Clean up modern pollen and climate dataframe to prepare for transfer function and reconstruction diagnostics. We have already changed the data to proportions, so just select the climate variables of interest.

# If you do not want to run the code for creating MPCT_proportions, then you can uncomment and run the following line:
# MPCT_proportions <- paleomat::MPCT_proportions

MPCT_proportions <-
  MPCT_climate_final3 %>%
  dplyr::mutate(
    climate =
      purrr::map(climate,
                 function(x) {
                   x %>%
                     dplyr::select(month, tavg)
                 })
  )

# If you need to write the data to file, then you can uncomment and run the following.
# save(MPCT_proportions, file = here::here("data/MPCT_proportions.rda"))



# Fossil Dataset

# If you do not want to run the code for creating fossil_pollen, then you can uncomment and run the following line:
# fossil_pollen <- paleomat::fossil_pollen

# Prepare fossil pollen dataset for transfer function and reconstruction diagnostics.
fossil_final$pollen <-
  purrr::map(fossil_final$pollen, function(x)
    tibble::column_to_rownames(x, "sample.id"))

fossil_pollen <- fossil_final

# If you need to write the data to file, then you can uncomment and run the following.
# save(fossil_pollen, file = here::here("data/fossil_pollen.rda"))



# Keep only taxa that are the same between the modern and fossil pollen dataset.
keep_cols <- intersect(colnames(dplyr::bind_rows(fossil_pollen$pollen) %>% dplyr::select(-Other)), colnames(dplyr::bind_rows(MPCT_proportions$pollen)))


fossil_pollen <- fossil_pollen %>%
  dplyr::mutate(pollen =
                  purrr::map(pollen,
                             function(x) {
                               x %>% 
                                 dplyr::select(dplyr::all_of(keep_cols))
                             }))

MPCT_proportions <- MPCT_proportions %>%
  dplyr::mutate(pollen =
                  purrr::map(pollen,
                             function(x) {
                               x %>% 
                                 dplyr::select(dplyr::all_of(keep_cols))
                             }))


# Put the modern pollen data into a dataframe and make the sample.id as the row names.
spp <- dplyr::bind_rows(MPCT_proportions$pollen) %>%
  dplyr::bind_cols(sample.id = MPCT_proportions$sample.id, .) %>%
  tibble::column_to_rownames("sample.id")

# Put the modern climate data into a dataframe and make the sample.id as the row names.
env <- MPCT_proportions %>%
  dplyr::select(sample.id, climate) %>%
  tidyr::unnest(climate) %>%
  # Pull just the climate (tavg) values as a vector.
  dplyr::pull(tavg)

# Turn climate value (tavg) vector into a named vector with the sample ids.
names(env) <- c(MPCT_proportions$sample.id)

```

# Transfer Function and Reconstruction Diagnostics and Reconstruction

Here, we use several transfer function and reconstruction diagnostics, such as squared residual length, the proportion of variance in the fossil data explained by an environmental reconstruction, and the no-analog threshold.

## Squared-residual Length

**Step 1**: We use squared residual length. Here, we return the residual length for each sample, and the 0.95 quantile.

```{r squared_residual_length, cache = TRUE, results = 'asis', warning = FALSE}

# If you do not want to run the code for creating rlen_results, then you can uncomment and run the following line:
# rlen_results <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/rlen_results.rds"))

rlen_results <-
  purrr::map(
    fossil_pollen$pollen,
    .f = function(z) {
      rlen <- analogue::residLen(spp, env, z, method = "cca")
      
      list(res_lengths = stack(rlen$passive),
           quantile_95 = as.numeric(quantile(rlen$train, probs = 0.95)))
    }
  )
# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   rlen_results,
#   here::here("vignettes/data/diagnostics_reconstruction/rlen_results.rds"))

# Extract the 0.95 quantile baseline to be used for sample removal.
rlen_ninety5quant <- rlen_results[[1]]$quantile_95

# Put the squared residual length and sample.id into one dataframe.
rlen_results <- purrr::map_dfr(rlen_results, `[[`, 1) %>%
  dplyr::rename("Squared_res_lengths" = 1, "sample.id" = 2)

# Extract the sample.ids for samples that will need to be removed since they did not meet the 0.95 threshold.
rlen_removal <-
  as.numeric(levels(rlen_results$sample.id))[rlen_results$Squared_res_lengths > rlen_ninety5quant]

# Results from rlen with sample.id for last 6000 years.
rlen_full_results1 <- rlen_results %>%
  dplyr::mutate(sample.id = as.integer(levels(sample.id))) %>%
  dplyr::filter(sample.id %in% fossil_west_post5500$sample.id)

# If you do not want to run the code for creating rlen_full_results2, then you can uncomment and run the following line:
# rlen_full_results2 <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/rlen_full_results2.rds"))

# Putting in the rlen results into a list structure.
rlen_full_results2 <-
  structure(list(res_lengths = rlen_full_results1,
                 quantile_95 = rlen_ninety5quant),
            class = "list")

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   rlen_full_results2,
#   here::here("vignettes/data/diagnostics_reconstruction/rlen_full_results2.rds"))

# Creating a copy of fossil pollen to remove samples that did not meet the 0.95 threshold from the fossil pollen dataset.
fos <- fossil_pollen
fos$pollen <-
  purrr::map(
    fos$pollen,
    ~ .x %>% tibble::rownames_to_column("sample.id") %>% dplyr::filter(!sample.id %in% rlen_removal) %>% tibble::column_to_rownames("sample.id")
  )

# Cleaning up fos to get the number of samples that are within each site. If a site is left with 0 or 1 samples, then it is removed.

# If you do not want to run the code for creating fos2, then you can uncomment and run the following line:
# fos2 <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/fos2.rds"))

# Remove any sites that have 0 samples from the fossil pollen dataset.
fos2 <- fos %>%
  dplyr::select(-n_samples) %>%
  dplyr::mutate(rows = purrr::map_dbl(pollen, ~ nrow(.x))) %>%
  dplyr::filter(!rows %in% c(0))

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(fos2,
#                  here::here("vignettes/data/diagnostics_reconstruction/fos2.rds"))

```

## randomTF Testing

You can explore the fossil pollen sites using the Random TF function to understand how much of the variance can be explained by tJuly. We do not use this step to remove any sites.

```{r randomTF, cache = TRUE, results = 'asis', warning = FALSE}

# # This line takes approximately 40 minutes to run (depending on your system). If you would like to re-run, then you can run the interior contents of the if statement. 
# if (!file.exists(here::here(
#   "vignettes/data/diagnostics_reconstruction/sig_results2_final.rds"
# ))) {
#   sig_results2_full <-
#     purrr::map2(fos2$pollen, fos2$dataset.id, function(x, y) {
#       output <- palaeoSig::randomTF(
#         spp = spp,
#         env = data.frame(tjul = env),
#         fos = x,
#         n = 10000,
#         fun = MAT,
#         col = 1
#       )
#       prop <- output$EX
#       names(prop) <- y
#       sig_out <- output$sig
#       names(sig_out) <- y
#       return(list(prop, sig_out))
#     })
#   
#   # Extract the proportion explained variance and the dataset ID.
#   sig_results2_proportions <- sapply(sig_results2_full, "[[", 1) %>%
#     stack(.) %>%
#     dplyr::rename(dataset.id = ind, proportion = values)
#   
#   # Extract the p-values and the dataset ID.
#   sig_results2_pvalues <- sapply(sig_results2_full, "[[", 2) %>%
#     stack(.) %>%
#     dplyr::rename(dataset.id = ind, pvalue = values)
#   
#   # Put the proportion of variance and p-value into the same dataframe.
#   sig_results2_final <-
#     dplyr::left_join(sig_results2_proportions, sig_results2_pvalues, by = "dataset.id") %>%
#     dplyr::select(dataset.id, proportion, pvalue) %>%
#     dplyr::mutate(dataset.id = as.numeric(levels(dataset.id)))
#   
#   readr::write_rds(
#     sig_results2_final,
#     here::here(
#       "vignettes/data/diagnostics_reconstruction/sig_results2_final.rds"
#     )
#   )
# }
# sig_results2_final <-
#   readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/sig_results2_final.rds"))
# 
# # Pull out the dataset IDs that have p-values less than the 0.10.
# only_90b <- sig_results2_final %>%
#   dplyr::filter(pvalue <= 0.10) %>%
#   dplyr::pull(dataset.id)
# 
# # Finally, subset the fossil pollen dataset by keeping only the dataset IDs that passed randomTF.
# fos3_90 <- fos %>%
#   dplyr::filter(dataset.id %in% only_90b)
# 

```

## MAT Transfer Function and Prediction

### Spatial Autocorrelation
```{r spatial_autocorrelation, cache = TRUE, results = 'asis', warning = FALSE}

# To reduce the effects of spatial autocorrelation, we first get the geographic euclidean distance between fossil and modern pollen sites. So, that we do not use samples within 300 km of the target site.

# Get list of modern pollen samples and geometry.
fos3_90 <- fos2
modern_samples <- MPCT_proportions %>% 
  dplyr::select(sample.id, geometry)

# Get list of fossil samples and geometry.
fossil_samples <- fos3_90 %>% 
  dplyr::mutate(pollen = purrr::map(
    pollen,
    ~ .x %>% tibble::rownames_to_column("sample.id") %>% dplyr::select(sample.id)
  )) %>% 
  tidyr::unnest(cols = "pollen") %>% 
  dplyr::select(sample.id, geometry)

# Calculate distances between all modern and fossil samples.
sample_dist <- sf::st_distance(modern_samples, fossil_samples) %>% 
  data.frame() %>% 
  dplyr::mutate(dplyr::across(dplyr::everything(), ~as.numeric(.x)/1000))

colnames(sample_dist) <- fossil_samples$sample.id
rownames(sample_dist) <- modern_samples$sample.id

```

**Step 2**: Run MAT, then do prediction/reconstruction.

```{r MAT_transfer_prediction, cache = TRUE, results = 'asis', warning = FALSE}

# Remove any columns in the modern pollen dataset that have 0 for the pollen taxa.
pollen_input <- spp %>%
  dplyr::select(which(colSums(.) > 0))

# Remove any columns in the fossil pollen dataset that have 0 for the pollen taxa.
fossil_input <- do.call(rbind, unname(fos3_90$pollen)) %>%
  dplyr::select(which(colSums(.) > 0))

# Keep only columns that are the same between the modern and fossil pollen dataset, as they need to be the same to run MAT.
keep_cols <- colnames(pollen_input[colSums(pollen_input) > 0])
keep_cols <- keep_cols[keep_cols %in% colnames(fossil_input)]

pollen_input <- pollen_input[, keep_cols]
fossil_input <- fossil_input[, keep_cols]

# Exclude modern samples if sample is within h distance to target (fossil pollen sample). Here, I get a list of the modern samples that are greater than 300 km for each site.
sample_dist2 <- sample_dist %>%
  tibble::rownames_to_column("modern.sample.id") %>%
  tidyr::pivot_longer(
    -modern.sample.id,
    names_to = "fossil.sample.id",
    values_to = "dist_km",
    values_drop_na = TRUE
  ) %>%
  dplyr::filter(dist_km > 300) %>%
  dplyr::select(-dist_km) %>%
  dplyr::group_by(fossil.sample.id) %>%
  dplyr::mutate(modern.sample.id = as.numeric(modern.sample.id)) %>%
  dplyr::summarize(modern.sample.id = list(unique(modern.sample.id))) %>%
  dplyr::mutate(fossil.sample.id = as.numeric(fossil.sample.id)) %>%
  dplyr::left_join(
    .,
    fossil_west_post5500 %>%
      dplyr::select(sample.id, site.id) %>%
      sf::st_drop_geometry(),
    by = c("fossil.sample.id" = "sample.id")
  ) %>%
  dplyr::select(-fossil.sample.id) %>%
  dplyr::distinct()

# Put fossil pollen data into a list where each site is its own dataframe.
fossil_input2 <- fossil_input %>%
  tibble::rownames_to_column("sample.id") %>%
  dplyr::mutate(sample.id = as.integer(sample.id)) %>%
  dplyr::left_join(
    .,
    fossil_west_post5500 %>%
      dplyr::select(sample.id, site.id) %>%
      sf::st_drop_geometry(),
    by = "sample.id"
  ) %>%
  dplyr::group_split(site.id) %>%
  purrr::set_names(purrr::map_chr(., ~ as.character(.x$site.id[1]))) %>%
  purrr::map(.,
             ~ .x %>%
               dplyr::select(-site.id) %>%
               tibble::column_to_rownames("sample.id"))


# If you want to run the code for creating reconstructions (i.e., the predict.mat and predict.mat bootstrap output for each site), then you can uncomment the section below (the code takes approximately 30 minutes to run, depending on your machine). Below, we load output files from the model to save time.

# You can use this line once you have ran the reconstruction. The file output was too large to load into GitHub.
# reconstructions <- readRDS(here::here("vignettes/data/full_reconstruction_output_FINAL.rds"))

# Iterate over the fossil pollen data to create reconstructions for each transfer function. 
# reconstructions <-
#   purrr::map2(.x = fossil_input2, .y = as.numeric(names(fossil_input2)), function(.x, .y) {
#     modern_samps <- sample_dist2 %>%
#       dplyr::filter(site.id == .y) %>%
#       dplyr::select(modern.sample.id) %>%
#       unlist() %>%
#       unname()
#     
#     pollen_input2 <- pollen_input %>%
#       tibble::rownames_to_column("sample.id") %>%
#       dplyr::mutate(sample.id = as.integer(sample.id)) %>%
#       dplyr::filter(sample.id %in% modern_samps) %>%
#       tibble::column_to_rownames("sample.id")
#     
#     env2 <- env[as.numeric(names(env)) %in% modern_samps]
#     
#     noK <- 4
#     # Create the transfer function using mat from the analogue package.
#     swap.mat2 <-
#       analogue::mat(pollen_input2, env2, method = "SQchord")
#     
#     # Create the transfer function using mat from the analogue package.
#     rlgh.mat2 <-
#       analogue:::predict.mat(object = swap.mat2,
#                              newdata = .x,
#                              k = noK)
#     
#     # Create the transfer function using mat from the analogue package. Bootstrapping takes approximately 5-10 minutes (depending on your system).
#     rlgh.mat2b4 <-
#       analogue:::predict.mat(
#         object = swap.mat2,
#         newdata = .x,
#         k = noK,
#         bootstrap = TRUE
#       )
#     
#     mat_prediction_4 <-
#       as.data.frame(rlgh.mat2b4$predictions$model$predicted[noK, ]) %>%
#       dplyr::mutate(site.id = as.numeric(.y)) %>%
#       tibble::rownames_to_column("sample.id") %>%
#       dplyr::select(sample.id, site.id, value = 2) %>%
#       # Pull out the bootstrap predicted values.
#       dplyr::left_join(
#         .,
#         as.data.frame(rlgh.mat2b4[["predictions"]][["bootstrap"]][["predicted"]][, noK]) %>%
#           dplyr::mutate(site.id = as.numeric(.y)) %>%
#           tibble::rownames_to_column("sample.id") %>%
#           dplyr::select(sample.id, site.id, bootstrap = 2),
#         by = c("sample.id", "site.id")
#       ) %>%
#       # Pull out the bootstrap rmsep.
#       dplyr::left_join(
#         .,
#         as.data.frame(rlgh.mat2b4[["predictions"]][["sample.errors"]][["rmsep"]][, noK]) %>%
#           dplyr::mutate(site.id = as.numeric(.y)) %>%
#           tibble::rownames_to_column("sample.id") %>%
#           dplyr::select(sample.id, site.id, rmsep = 2),
#         by = c("sample.id", "site.id")
#       ) %>%
#       # Pull out the bootstrap s1 errors.
#       dplyr::left_join(
#         .,
#         as.data.frame(rlgh.mat2b4[["predictions"]][["sample.errors"]][["s1"]][, noK]) %>%
#           dplyr::mutate(site.id = as.numeric(.y)) %>%
#           tibble::rownames_to_column("sample.id") %>%
#           dplyr::select(sample.id, site.id, s1 = 2),
#         by = c("sample.id", "site.id")
#       )
#     
#     return(
#       list(
#         fossil.sample.id = .y,
#         recon = rlgh.mat2,
#         recon_bootstrap = rlgh.mat2b4,
#         recon_df = mat_prediction_4
#       )
#     )
#     
#   })

# If you need to write the data to file, then you can uncomment and run the following.
# saveRDS(reconstructions, here::here("vignettes/data/full_reconstruction_output_FINAL.rds"))

# Save output of MAT.
# If you do not want to run the code for creating mat_transfer, then you can uncomment and run the following line:
# mat_transfer <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/mat_calibration.rds"))

# Create the transfer function using mat from the analogue package.
mat_transfer <- analogue::mat(pollen_input, env, method = "SQchord")
noK <- 4

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   mat_transfer,
#   here::here("vignettes/data/diagnostics_reconstruction/mat_calibration.rds"))

# Run the prediction. If you do not want to run the code for creating mat_predict, then you can uncomment and run the following line:
# mat_predict <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/mat_prediction.rds"))

# Create the transfer function using mat from the analogue package.
mat_predict <-
  analogue:::predict.mat(object = mat_transfer,
                         newdata = fossil_input,
                         k = noK)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   mat_predict,
#   here::here("vignettes/data/diagnostics_reconstruction/mat_prediction.rds"))



# Here, we pull out the data from the predict.mat objects. First, we pull out the predicted values for the four best analogues. In the analogue package, the analogues have cumulative means, so we only need to extract the four record here.
# Since, I do not run the reconstructions script here. I saved the output as a tibble.
mat_prediction_4 <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/mat_prediction_4.rds"))

# mat_prediction_4 <-
#   purrr::map_df(reconstructions, "recon_df")

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   mat_prediction_4,
#   here::here("vignettes/data/diagnostics_reconstruction/mat_prediction_4.rds"))


# Since, I do not run the reconstructions script here. I saved the output as a tibble.
predictMat <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/predictMat.rds"))
# predictMat <- purrr::map(reconstructions, "recon")

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   predictMat,
#   here::here("vignettes/data/diagnostics_reconstruction/predictMat.rds"))

```


## No Analog Threshold (or Minimum Dissimilarity)

**Step 3**: Remove any fossil pollen samples that have a minimum dissimilarity that is greater than the 5% quantile.

```{r no_analog, cache = TRUE, results = 'asis', warning = FALSE}

# If you do not want to run the code for creating minDC_full_results, then you can uncomment and run the following line:
# minDC_full_results <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/minDC_full_results.rds"))

# Get a list of the samples that remain in the fossil pollen dataset after rlen and the randomTF removal steps.
pre_minDC_samples <- rlen_full_results1 %>%
  dplyr::filter(!sample.id %in% rlen_removal) %>%
  dplyr::left_join(
    .,
    fossil_west_post5500 %>%
      dplyr::select(dataset.id, site.id, sample.id),
    by = "sample.id"
  ) %>%
  #dplyr::filter(dataset.id %in% only_90b) %>%
  dplyr::pull(sample.id)

rlgh.mdc2 <- predictMat %>% 
  purrr::map_df(., "minDC") %>% 
  tidyr::pivot_longer(everything(), names_to = "ind", values_to = "values", values_drop_na = TRUE)

# Put the results into a list.
minDC_results <- rlgh.mdc2 %>%
  dplyr::mutate(ind = as.integer(ind)) %>%
  dplyr::filter(ind %in% pre_minDC_samples) %>%
  dplyr::left_join(
    .,
    fossil_west_post5500 %>%
      dplyr::select(sample.id, dataset.id, site.id, age) %>%
      dplyr::mutate(age = paleomat::BPtoBCAD(age)) %>%
      sf::st_drop_geometry(),
    by = c("ind"  = "sample.id")
  )

# Put the results into a list structure.
minDC_full_results <- structure(list(minDC = minDC_results,
                                     quantiles = mat_predict$quantiles),
                                class = "list")

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   minDC_full_results,
#   here::here("vignettes/data/diagnostics_reconstruction/minDC_full_results.rds"))



# Put data into a list. Create list of samples that did and did not meet the threshold, and subset the fossil pollen data to the samples that did pass the minDC test.

# If you do not want to run the code for creating minDC_summary, then you can uncomment and run the following line:
# minDC_summary <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/minDC_summary.rds"))

# Get sample IDs and dissimilarity for samples that did not pass the threshold (i.e., samples that will need to be removed).
no_analog_threshold <- mat_predict[["quantiles"]][["5%"]]

remove_samples <- rlgh.mdc2 %>%
  dplyr::rename(dissimilarity = 2, sample.id = ind) %>%
  dplyr::filter(dissimilarity > no_analog_threshold)

# Remove the samples from the dataset.
mat_prediction_cleaned <- mat_prediction_4 %>%
  dplyr::filter(!sample.id %in% remove_samples$sample.id)

# Final Sample set.
final_approved_samples <- minDC_full_results$minDC %>%
  dplyr::filter(!ind %in% remove_samples$sample.id) %>%
  dplyr::pull(ind)

# Put into a list structure.
minDC_summary <- list(
  remove_samples = remove_samples,
  mat_prediction_cleaned = mat_prediction_cleaned,
  final_approved_samples = final_approved_samples
)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   minDC_summary,
#   here::here("vignettes/data/diagnostics_reconstruction/minDC_summary.rds"))

```

## Clean Up Data

Clean up fossil pollen data to create the final dataset of samples.

```{r clean_up_data, cache = TRUE, results = 'asis', warning = FALSE}

# If you do not want to run the code for creating fos_90, then you can uncomment and run the following line:
# fos_90 <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/fos_90.rds"))

fos_90 <- mat_prediction_cleaned %>%
  dplyr::mutate(sample.id = as.integer(sample.id)) %>%
  dplyr::filter(sample.id %in% final_approved_samples) %>%
  dplyr::left_join(.,
                   fossil_west_post5500 %>%
                     dplyr::select(dataset.id:pub_year),
                   by = "sample.id") %>%
  dplyr::filter(age <= 6000) %>%
  dplyr::mutate(age = round(age)) %>%
  sf::st_as_sf()

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(fos_90,
#                  here::here("vignettes/data/diagnostics_reconstruction/fos_90.rds"))

```

# Anomaly Calculation

```{r anomaly_calculation, cache = TRUE, results = 'asis', warning = FALSE}

# Extract the modern mean temperature at each fossil pollen site, and then calculate the anomalies.

# If you do not want to run the code for creating fos_90, then you can uncomment and run the following line:
# site.preds.unlisted.anom <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/final_paleomat_dataset3.rds"))

# Put lat and long into individual columns.
tavg_new_90 <- fos_90 %>%
  dplyr::mutate(long = unlist(purrr::map(.$geometry, 1)),
                lat = unlist(purrr::map(.$geometry, 2))) %>%
  # Remove/drop geometry. 
  dplyr::select(-geometry) %>%
  sf::st_drop_geometry() %>%
  dplyr::mutate(age = mean) %>% 
  # Create a date column with BC/AD ages.
  dplyr::mutate(date = paleomat::BPtoBCAD(mean)) %>%
  # Remove any data older than 5500 years old.
  dplyr::filter(age <= 5500) %>%
  # Create 100 year periods. 
  dplyr::group_by(period = cut(date, breaks = seq(-5500, 2100, 100))) %>%
  # Get a midpoint for each period.
  dplyr::mutate(mid_period = paleomat::get_midpoints(period))

# Extract the modern mean July temperature (PRISM data) from each fossil pollen site. 
modern_temperature <-
  as.data.frame(
    raster::extract(
      x =  paleomat::temp.raster,
      y = tavg_new_90 %>% dplyr::ungroup() %>%
        dplyr::select(long, lat)
    )
  ) %>%
  dplyr::rename(modern = 1)

# Bind the modern data to fossil pollen dataset.
site.preds.unlisted <-
  dplyr::bind_cols(tavg_new_90, modern = modern_temperature$modern)

# Calculate the temperature anomaly and the bootstrapped temperature anomaly.
site.preds.unlisted.anom <- site.preds.unlisted %>% 
  dplyr::rowwise() %>%
  dplyr::mutate(anom = value - modern) %>%
  dplyr::mutate(anom_bootstrap = bootstrap - modern)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   site.preds.unlisted.anom,
#   here::here("vignettes/data/diagnostics_reconstruction/final_paleomat_dataset3.rds"))

knitr::kable(site.preds.unlisted.anom,
             caption = "Reconstruction results from MAT with calculated anomalies, RMSEP, leave-one-out errors, and metadata.")

```

# Age-depth Simulation

```{r age_depth_simulation, cache = TRUE, results = 'asis', warning = FALSE}

# If you do not want to run the code for creating final_paleomat_output, then you can uncomment and run the following line:
# final_paleomat_output <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/final_paleomat_output.rds"))

site.preds.unlisted.anom2 <- site.preds.unlisted.anom
# Initialize a Year column. 
site.preds.unlisted.anom2$Year <- NA

site.preds.unlisted.anom2 <- site.preds.unlisted.anom2 %>%
  dplyr::group_by(dataset.id) %>% 
  tidyr::nest()

final_paleomat_output <- site.preds.unlisted.anom2 %>%
  dplyr::mutate(data = purrr::map(
    data,
    ~ purrr::map_dfr(
      # We run 100 simulations for the age-depth model.
      1:100,
      \(i) paleomat::simulate_age_model(.) %>% dplyr::mutate(Simulation = i, Year = paleomat::BPtoBCAD(Year))
    )
  )) %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::bind_rows(.,
                   site.preds.unlisted.anom %>%
                     dplyr::mutate(Year = paleomat::BPtoBCAD(mean),
                                   Simulation = 0)) %>%
  dplyr::mutate(date = Year)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   final_paleomat_output,
#   here::here("vignettes/data/diagnostics_reconstruction/final_paleomat_output.rds"))


# Getting ages for all 100 simulations for all fossil pollen samples (n = 773). This is for plotting in the supplemental figures.
# If you do not want to run the code for creating sim_ages3, then you can uncomment and run the following line:
# sim_ages3 <- readr::read_rds(here::here("vignettes/data/diagnostics_reconstruction/sim_ages3.rds"))
sim_ages <- paleomat::fossil_west_post5500 %>% 
  dplyr::select(dataset.id:mean)

sim_ages2 <- sim_ages

sim_ages2$Year <- NA

sim_ages2 <- sim_ages2 %>%
  dplyr::select(dataset.id:mean, Year) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::group_by(dataset.id) %>% 
  tidyr::nest()

sim_ages3 <- sim_ages2 %>%
  dplyr::mutate(data = purrr::map(
    data,
    ~ purrr::map_dfr(
      1:100,
      \(i) paleomat::simulate_age_model(.) %>% dplyr::mutate(Simulation = i, Year = paleomat::BPtoBCAD(Year))
    )
  )) %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::bind_rows(.,
                   sim_ages %>%
                     dplyr::mutate(Year = paleomat::BPtoBCAD(mean),
                                   Simulation = 0)) %>%
  dplyr::mutate(date = Year)

# If you need to write the data to file, then you can uncomment and run the following.
# readr::write_rds(
#   sim_ages3,
#   here::here("vignettes/data/diagnostics_reconstruction/sim_ages3.rds"))

```

For figures and tables in the journal article, please see "Paleomat_Figures.Rmd".


```{r sample_numbers_for_article, cache = TRUE, results = 'asis', warning = FALSE}

# Get counts for throughout this Rmarkdown to show how number of sites, datasets, and samples change.

# Get elevation stats for fossil pollen sites.
fossil_elev <- fossil_metadata_counts_final3 %>% 
  dplyr::select(site.id) %>% 
  dplyr::distinct() %>% 
  elevatr::get_elev_point(., prj = 4326, src = "aws")

fossil_elev2 <- fossil_elev %>% 
  dplyr::filter(site.id %in% fossil_metadata_counts_final5$site.id)

fossil_elev3 <- fossil_elev %>% 
  dplyr::filter(site.id %in% paleomat::fossil_west_post5500$site.id)

fos_join <- dplyr::left_join(fos2, fossil_metadata_counts_final5 %>% dplyr::select(site.id, dataset.id) %>% sf::st_drop_geometry() %>% dplyr::distinct(), by = "dataset.id")

# Put all information into a dataframe.
data.frame(Stage = c("Initial", "Quality Control", "Less Than 6000 BP", "resLength", "No Analogue"),
          noSites = c(length(unique(fossil_metadata_counts_final3$site.id)),
                      length(unique(fossil_metadata_counts_final5$site.id)),
                      length(unique(fossil_west_post5500$site.id)),
                      length(unique(fos_join$site.id)),
                      length(unique(fos_90$site.id.x))
                      ),
          noDatasets = c(length(unique(fossil_metadata_counts_final3$dataset.id)),
                         length(unique(fossil_metadata_counts_final5$dataset.id)),
                         length(unique(fossil_west_post5500$dataset.id)),
                         length(unique(fos_join$dataset.id)),
                         length(unique(fos_90$dataset.id))
                         ),
          noSamples = c(nrow(fossil_metadata_counts_final3),
                        nrow(fossil_metadata_counts_final5),
                        nrow(fossil_west_post5500),
                        nrow(dplyr::bind_rows(fos_join$pollen)),
                        nrow(fos_90)
                        ),
          meanElev = c(mean(fossil_elev$elevation),
                       mean(fossil_elev2$elevation),
                       mean(fossil_elev3$elevation),
                       NA,
                       NA),
          minElev = c(min(fossil_elev$elevation),
                      min(fossil_elev2$elevation),
                      min(fossil_elev3$elevation),
                      NA,
                      NA),
          maxElev = c(max(fossil_elev$elevation),
                      max(fossil_elev2$elevation),
                      max(fossil_elev3$elevation),
                      NA,
                      NA)
          )

```
